---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---

## set up docker container: run this in the powershell
docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop


##new function to read in results with validation option
This would be run locally to look at outputs
```{r}

##function to read in and plot quickly
#libraries
library(RHESSysIOinR)
library(tidyverse)
#unit conversion functions
source("../read_plot_rhessys_fun.R")

#read in validation csv
target_df <- read_csv("../Target_Data_comp/TargetData_15mar26.csv") |> 
  rename(date = Date)

```



## run model using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

All the code below would be run in the docker after opening with the line at the top of the script

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NewMaps")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2300 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinuptests"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/world_flow_kmeans300_patches/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_tests",
  flowtable        = "worldfiles/world_flow_kmeans300_patches/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Ntests",
  output_prefix    = "HPB_spinup350yr",
  commandline_options = "-b -g -climrepeat"  
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro params
  list(input_hdr$soil_def[1], "m",                  c(0.1, 0.1, 0.1, 0.1)),
  list(input_hdr$soil_def[1], "Ksat_0",             c(404, 404,  404, 404)),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0.1, 0.1, 0, 0)),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0.9, 0.9, 0, 0)),
  
  ##set soil params
  list(input_hdr$soil_def[1], "N_decay", c(3,3,3,3)), #
  list(input_hdr$soil_def[1], "NO3_adsorption_rate", c(0.01,0.01,0.01,0.01)), #0 is default
  list(input_hdr$soil_def[1], "DON_production_rate", c(0.015,0.015,0.015,0.015)), #

  
  ## set veg params
  list(input_hdr$stratum_def[1], "epc.max_lai", c(6,6,6,6)), #def is 20
  list(input_hdr$stratum_def[1], "epc.livewood_cn", c(200,200,200,200)), #def is 48
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc", c(1.7,1.7,1.7,1.7)), #
  list(input_hdr$stratum_def[1], "epc.leaf_cn", c(2.2,2.2,2.2,2.2)), #
  list(input_hdr$stratum_def[1], "epc.alloc_frootc_leafc", c(0.9,0.9,0.9,0.9)), #
  list(input_hdr$stratum_def[1], "epc.branch_turnover", c(0.001, 0.001, 0.001, 0.001)) #

  # list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)) #1 is default
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
#   def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output
#   return_cmd    = F   # flip to FALSE to run
# )


#run for multi param
run_rhessys_multi(
  input_rhessys   = input_rhessys,
  hdr_files       = input_hdr,
  tec_data        = input_tec_data,
  def_pars        = input_def_pars,
  output_filter   = NULL,
  all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
  return_cmd      = F
)


```


## Look at first spin output

```{r}
zz <- readin_rhessys_output(pre = "out/Ntests/HPB_spinup350yr_run1")
head(zz$bdg)

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run1", date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out")  ) # NULL is default

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run2",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out") ) 

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run3",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out") ) 

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run4",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out") ) 



# read_plot_rhessys("out/Spin/Spin_valid2/HPB_spinup450yr", validate = TRUE)


```





## Run for a few years off the tests

```{r}
# #Can run all of this within the Docker container after opening R
# 
# setwd("repo/HPB_files_NEW")
# library(RHESSysIOinR)
# 
# dates = c("2015 1 1 1", "2026 1 2 24") #has to end with a 24 hour so that the tec output will work
# 
# name = "hpb_spinup_Hydro"
# 
# input_rhessys = IOin_rhessys_input(
#   version          = "../../rhessys7.5",
#   tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
#   world_file       = "worldfiles/NO3GWtests/hpb.world4.Y2300M1D2H23.state", #what was made in pre-processing
#   world_hdr_prefix = "hpb_spintest",
#   flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
#   start            = dates[1],
#   end              = dates[2],
#   output_folder    = "out/NO3_GW_tests/ValidRun4",
#   output_prefix    = "HPB_spinup250yr",
#   commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
# )
# 
# #header directly copied from hpb.hdr
# input_hdr = IOin_hdr(
#   basin      = "defs/basin.def",
#   hillslope  = "defs/hill.def",
#   zone       = "defs/zone.def",
#   soil       = "defs/soil_siltyloam.def",
#   landuse    = "defs/lu_undev.def",
#   stratum    = "defs/veg_deciduous.def",
#   basestations = "clim/ccr_base"
# )
# 
# #set param list
# pars_list = list(
#   list(input_hdr$soil_def[1], "m",                  c(0.108)),
#   list(input_hdr$soil_def[1], "Ksat_0",             c(404.3)),
#   list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0)), #c(0.1, 0.1, 0, 0)
#   list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0)), #c(0.9, 0.9, 0, 0)
#   list(input_hdr$soil_def[1], "NO3_adsorption_rate", c(0.9)) #c(0.1, 0.9, 0.1, 0.9)    #0 is default
#     )
# 
# # #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
# input_def_pars = IOin_def_pars_simple(pars_list)
# 
# 
# #make tec file locally
# input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = F)
# 
# 
# #run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
#   def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output
#   return_cmd    = F   # flip to FALSE to run
# )


```


```{r}
read_plot_rhessys("out/NO3_GW_tests/ValidRun1/HPB_spinup250yr", validate = TRUE)


```








