---
title: "HPB_preprocess"
output: html_document
date: "2026-04-06"
---


## Get packages 

```{r}
library(terra)
library(RHESSysPreprocessing)
```



#### look at maps 

```{r}
#basin
getwd()
basin <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_ws.tif")
plot(basin, main = "Watershed")

#dem
dem <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_dem.tif")
plot(dem, main = "DEM")

#slope
slope <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_slope.tif")
plot(slope, main = "slope")

#aspect
aspect <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_aspect.tif")
plot(aspect, main = "aspect")

#West Horizon
whr <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_whr.tif")
plot(whr, main = "West horizon")

#East horizon
ehr <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_ehr.tif")
plot(ehr, main = "east  horizon")

#slope
slope <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_slope.tif")
plot(dem, main = "slope")



#basins
basins <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_basin750.tif")
plot(basins, main = "basins")

# half basins
halfbasins <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_halfbasin750.tif")
plot(halfbasins, main = "halfbasins")


#streams Grass default 
streams <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_stream750.tif")
plot(streams, main = "streams", type = "classes") #the type classes let you see what the segment numbers are


#streams as 1
streams_1 <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_stream750_1.tif")
plot(streams_1, main = "streams_1", type = "classes")


```


#### Try running RHESSys preprocess with these files from GRASS


```{r}

getwd()
## set up file paths 
template = "worldfiles/templates/template_hpb_NOfill.txt"
type = "raster"
typepars = "spatial_data" #path of the folder containing your rasters; these are copied from grass output in "../Spatial_Data/GRASS_hpbNEW_outputs/"
name = "hpb"
overwrite = FALSE
streams = "hpb_stream750"  #this is the GRASS streams that has segment #s
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)


```


#fill NAs for error sites 
basin750, halfbasin 750, slope, aspect

```{r}

#read in rasters and check for visual NA issues
watershed <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_ws.tif")
plot((watershed), main = "watershed map")
plot(is.na(watershed), main = "NA locations in watershed map")

#slope and aspect map with NA
slope <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_slope.tif")
plot(slope, main = "slope map") # type shows the legend
plot(is.na(slope), main = "NA locations in slope map")

aspect <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_aspect.tif")
plot(aspect, main = "aspect map") # type shows the legend
plot(is.na(aspect), main = "NA locations in aspect map")

#basin and patches with NA
basins <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_basin750.tif")
unique(basins)
plot(basins, type = 'classes', main = "basins map") # type shows the legend
plot(is.na(basins), main = "NA locations in basins map")

halfbasins <- rast("../Spatial_Data/GRASS_hpbNEW_outputs/hpb_halfbasin750.tif")
unique(halfbasins)
plot(halfbasins, type = 'classes', main = "halfbasins map") # type shows the legend
plot(is.na(halfbasins), main = "NA locations in halfbasins map")


#check that extent and everything is the same, should return trues
compareGeom(basins, watershed)
compareGeom(halfbasins, watershed)
compareGeom(slope, watershed)
compareGeom(aspect, watershed)

#visualize NAs better
na_mask <- is.na(basins)
plot(na_mask, col = c("transparent", "red"), legend = FALSE, main = "NA Locations")

# zoom(patch)

```


```{r}
#function from chatt
fill_edge_na <- function(r, watershed, maxiter = 20) {
  
  # Ensure aligned
  r <- resample(r, watershed, method="near")
  
  # Mask to watershed
  r <- mask(r, watershed)
  
  for (i in 1:maxiter) {
    
    na_cells <- is.na(r) & !is.na(watershed)
    if (!any(values(na_cells), na.rm=TRUE)) break
    
    r <- focal(
      r,
      w = 3,
      fun = function(x, ...) {
        if (all(is.na(x))) return(NA)
        return(x[which(!is.na(x))[1]])
      },
      na.policy = "only",
      fillvalue = NA
    )
    
    r <- mask(r, watershed)
  }
  
  return(r)
}


#### try it 
#basins
basins_filled  <- fill_edge_na(basins, watershed)
plot(basins_filled)
sum(is.na(values(basins_filled)))
sum(is.na(values(basins)))
sum(is.na(values(watershed)))

#half basisn
halfbasins_filled <- fill_edge_na(halfbasins, watershed)
plot(halfbasins_filled)
sum(is.na(values(halfbasins_filled)))
sum(is.na(values(halfbasins)))
sum(is.na(values(watershed)))


##write new rasters
# writeRaster(basins_filled, "spatial_data/hpb_basin750_filled_6apr26.tif", overwrite = T)
# writeRaster(halfbasins_filled, "spatial_data/hpb_halfbasin750_filled_6apr26.tif", overwrite = T)


#slope 
slope_filled  <- fill_edge_na(slope, watershed)
plot(slope)
plot(slope_filled)
sum(is.na(values(slope_filled)))
sum(is.na(values(slope)))
sum(is.na(values(watershed)))

#aspect
aspect_filled  <- fill_edge_na(aspect, watershed)
plot(aspect)
plot(aspect_filled)
sum(is.na(values(aspect_filled)))
sum(is.na(values(aspect)))
sum(is.na(values(watershed)))

##write new rasters
# writeRaster(slope_filled, "spatial_data/hpb_slope_filled_6apr26.tif", overwrite = T)
# writeRaster(aspect_filled, "spatial_data/hpb_aspect_filled_6apr26.tif", overwrite = T)



```



## Retry preprocess scripts with fill

```{r}

getwd()
## set up file paths 
template = "worldfiles/templates/template_hpb_fill_halfbasinPatch.txt"
type = "raster"
typepars = "spatial_data" #path of the folder containing your rasters; these are copied from grass output in "../Spatial_Data/GRASS_hpbNEW_outputs/"
name = "hpb"
overwrite = FALSE
streams = "hpb_stream750"  #this is the GRASS streams that has segment #s ; doens't seem to affect flow table or worldfie using stream with segments or all segments as 1
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)

```



## some plots to conceputalize basins 

```{r}
#basins and their streams
plot(rast("./spatial_data/hpb_basin750_filled_6apr26.tif"))
plot(rast("./spatial_data/hpb_stream750.tif"), type = "classes", add = T)
plot(rast("./spatial_data/hpb_stream750.tif"), type = "classes", add = T,
          col  = c("steelblue", "orange", "forestgreen")) # one color per class

# col  = hcl.colors(n = 5, palette = "Set 2"),   # n = number of classes


#half basins
plot(rast("./spatial_data/hpb_halfbasin750_filled_6apr26.tif"))



```


## try making a more informed patch map 
Functionally doing the the altitude bands × hillslope method from the RHESSys wiki in GRASS, code adapted with claude

```{r}
dem  <- rast("./spatial_data/hpb_dem.tif")
plot(dem)

hillslopes <- rast("./spatial_data/hpb_halfbasin750_filled_6apr26.tif")
plot(hillslopes)

# ~100 patches across 6 hillslopes = ~17 elevation bands per hillslope
n_bands   <- round(100 / global(hillslopes, "max", na.rm = TRUE)[[1]])
elev_min  <- global(dem, "min", na.rm = TRUE)[[1]]
elev_range <- global(dem, "range", na.rm = TRUE)
band_width <- (elev_range[1,2] - elev_range[1,1]) / n_bands

# Create elevation bands
alt_bands <- floor((dem - elev_min) / band_width)
plot(alt_bands)

# Combine: each hillslope x elevation band = unique patch
patch_map <- hillslopes * (n_bands + 1) + alt_bands
patch_map <- mask(patch_map, hillslopes)
plot(patch_map)

# Compact IDs to 1:N with no gaps
vals <- values(patch_map)
uvals <- sort(unique(vals[!is.na(vals)]))
values(patch_map)[!is.na(vals)] <- match(vals[!is.na(vals)], uvals)

#look at patches a few ways
plot(patch_map)
plot(patch_map, type = "classes")

pal <- colorRampPalette(c("darkblue", "white", "darkred"))
# Use 100 shades for a smooth gradient
plot(patch_map, col = pal(100))

#check number of patches
unique(patch_map)
global(patch_map, "max", na.rm = TRUE)  # check total patches

##write raster
# writeRaster(patch_map, "./spatial_data/hpb_patch_map.tif", overwrite = TRUE)

plot(rast("./spatial_data/hpb_patch_map.tif""))

# Adjust the target by changing 100 — the actual count may differ slightly since not every hillslope spans the full elevation range. Check with global(patch_map, "max", na.rm = TRUE) and tune from there.


```



#run preprocessing with smaller patch sizes

```{r}

getwd()
## set up file paths 
template = "worldfiles/templates/template_hpb.txt"
type = "raster"
typepars = "spatial_data" #path of the folder containing your rasters; these are copied from grass output in "../Spatial_Data/GRASS_hpbNEW_outputs/"
name = "hpb"
overwrite = FALSE
streams = "hpb_stream750"  #this is the GRASS streams that has segment #s ; doens't seem to affect flow table or worldfie using stream with segments or all segments as 1
header = FALSE
parallel = TRUE
make_stream = 4

## run preprocess
RHESSysPreprocess(
  template,
  name,
  map_dir = typepars,
  typepars = NULL,
  streams = streams,
  overwrite = overwrite,
  roads = NULL,
  impervious = NULL,
  roofs = NULL,
  header = header,
  meta = FALSE,
  asprules = NULL,
  unique_strata_ID = TRUE,
  seq_patch_IDs = FALSE,
  output_patch_map = FALSE,
  fire_grid_out = FALSE,
  parallel = parallel,
  make_stream = make_stream,
  skip_hillslope_check = TRUE,
  convert_aspect = TRUE,
  wrapper = TRUE
)


```






