---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---

## set up docker container: run this in the powershell
docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop


##new function to read in results with validation option
This would be run locally to look at outputs
```{r}

##function to read in and plot quickly
#libraries
library(RHESSysIOinR)
library(tidyverse)
#unit conversion functions
source("../read_plot_rhessys_fun.R")

#read in validation csv
target_df <- read_csv("../Target_Data_comp/TargetData_15mar26.csv") |> 
  rename(date = Date)

```


#coweta param test for quinn

```{r}


setwd("repo/HPB_files_NewMaps")
#devtools::install_github("RHESSys/RHESSysIOinR")
library(RHESSysIOinR)


#### Run Spinup ####
dates = c("1950 1 1 1", "2350 1 2 24") #has to end with a 24 hour so that the tec output will work


name = "hpb_spin_cowtest"

output_folder = "out/apr13RouteTest/COWTest"
if (!dir.exists(output_folder)) dir.create(output_folder, recursive = TRUE)


input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/world_flow_kmeans300_patches/hpb.world", #what was made in pre-processing
  world_hdr_prefix = name,
  flowtable        = "worldfiles/world_flow_kmeans300_patches/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = output_folder,
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat" #if adding -p add before -g to get patch grow
  #commandline_options = "-b -g -climrepeat -str worldfiles/world_flow_kmeans300_patches/stream.hpb -stro"
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/cow_zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/cow_veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1), #0.1; 0.05
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9), #0.9, 0.2

  list(input_hdr$stratum_def[1], "mrc.per_N", 1),
  list(input_hdr$stratum_def[1], "epc.max_lai", 10) ,
  list(input_hdr$stratum_def[1], "kfrag_base", 0.0002 ) 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)





########################################
#### Run 5 years off of spinup ####
dates = c("2020 1 1 1", "2025 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_test_cowtest"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/world_flow_kmeans300_patches/hpb.world.Y2350M1D2H23.state", #what was made in pre-processing
  world_hdr_prefix = name,
  flowtable        = "worldfiles/world_flow_kmeans300_patches/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = output_folder,
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat" #if adding -p add before -g to get patch grow
  #commandline_options = "-b -g -climrepeat -str worldfiles/world_flow_kmeans300_patches/stream.hpb -stro"
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/cow_zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/cow_veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = F)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)


```


eval

```{r}
getwd()
read_plot_rhessys(outname = "out/apr13RouteTest/COWTest/hpb_spin_cowtest", 
                  date_filter = as.Date("1960-01-01"), ncol = 5, validate = F)


## validate
read_plot_rhessys(outname = "out/apr13RouteTest/COWTest/hpb_test_cowtest", ncol = 5, validate = T)

```

```{r}
#n-cycle validate
zz <- readin_rhessys_output("out/apr13RouteTest/COWTest/hpb_spin_cowtest")

zz$bdg |> dplyr::select(date, #year,
                        gwNO3out, streamflow_NO3
                        # N_dep, nuptake, nfix, nitrif, denitrif, #rates
                        # gwDONout, gwNH4out, gwNO3out #fluxes
                        # #stocks
                        # ,plantn, soiln,
                        # nitrate, sminn, surfaceN, npool, litrn,
                        # gwDON, gwNH4, gwNO3, streamflow_NO3, streamflow_NH4, streamflow_DON,
                        # StreamNO3_from_surface, StreamNO3_from_sub
                        ) |> 
  mutate(stream_gwNO3out = streamflow_NO3 - gwNO3out) |> 
  pivot_longer(-1) |> 
  ggplot(aes(x = date, y = value))+
  geom_point()+
  facet_wrap(~name, scales = "free_y")

```








##try using coweeta def files

```{r}

#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NewMaps")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2300 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_cowparam"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/world_flow_kmeans300_patches/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_tests",
  flowtable        = "worldfiles/world_flow_kmeans300_patches/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Coweeta_Params",
  output_prefix    = "HPB_Cow_350yr",
  commandline_options = "-b -g -climrepeat"  
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/cow_zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/cow_veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  list(input_hdr$soil_def[1], "m",                  c(0.1)),
  list(input_hdr$soil_def[1], "Ksat_0",             c(404)),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0.05)),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0.2))
  )


input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)

```


```{r}

read_plot_rhessys(outname = "out/Coweeta_Params/HPB_Cow_350yr", date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3")  ) # NULL is default


```




## run model using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

All the code below would be run in the docker after opening with the line at the top of the script

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NewMaps")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2300 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinuptests"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/world_flow_kmeans300_patches/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_tests",
  flowtable        = "worldfiles/world_flow_kmeans300_patches/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Ntests",
  output_prefix    = "HPB_spinup350yr2",
  commandline_options = "-b -g -climrepeat"  
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  list(input_hdr$soil_def[1], "m",                  c(0.1)),
  list(input_hdr$soil_def[1], "Ksat_0",             c(404)),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0.05)),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0.2)),
  
  ##set soil params
  # list(input_hdr$soil_def[1], "N_decay", c(3,3,3,3)), #
  list(input_hdr$soil_def[1], "NO3_adsorption_rate", c(0.001)), #0 is default
  list(input_hdr$soil_def[1], "active_zone_z", c(2)), #

  
  ## set veg params
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", c(5)), #def is 20
  list(input_hdr$stratum_def[1], "epc.livewood_cn", c(200)) #def is 48


  
  
  ## set hydro params
  # list(input_hdr$soil_def[1], "m",                  c(0.1, 0.1, 0.1, 0.1)),
  # list(input_hdr$soil_def[1], "Ksat_0",             c(404, 404,  404, 404)),
  # list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0.1, 0.1, 0, 0)),
  # list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0.9, 0.9, 0, 0)),
  # 
  # ##set soil params
  # list(input_hdr$soil_def[1], "N_decay", c(3,3,3,3)), #
  # list(input_hdr$soil_def[1], "NO3_adsorption_rate", c(0.01,0.01,0.01,0.01)), #0 is default
  # list(input_hdr$soil_def[1], "DON_production_rate", c(0.015,0.015,0.015,0.015)), #
  # 
  # 
  # ## set veg params
  # list(input_hdr$stratum_def[1], "epc.max_lai", c(6,6,6,6)), #def is 20
  # list(input_hdr$stratum_def[1], "epc.livewood_cn", c(200,200,200,200)), #def is 48
  # list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc", c(1.7,1.7,1.7,1.7)), #
  # list(input_hdr$stratum_def[1], "epc.leaf_cn", c(2.2,2.2,2.2,2.2)), #
  # list(input_hdr$stratum_def[1], "epc.alloc_frootc_leafc", c(0.9,0.9,0.9,0.9)), #
  # list(input_hdr$stratum_def[1], "epc.branch_turnover", c(0.001, 0.001, 0.001, 0.001)) #

  # list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)) #1 is default
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)


#run for multi param
# run_rhessys_multi(
#   input_rhessys   = input_rhessys,
#   hdr_files       = input_hdr,
#   tec_data        = input_tec_data,
#   def_pars        = input_def_pars,
#   output_filter   = NULL,
#   all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
#   return_cmd      = F
# )


```


##trying second round of claude param

```{r}

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr2", date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3")  ) # NULL is default


```




## Look at first spin output



```{r}
zz <- readin_rhessys_output(pre = "out/Ntests/HPB_spinup350yr_run1")
head(zz$bdg)

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run1", date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3")  ) # NULL is default

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run2",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3") ) 

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run3",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3") ) 

read_plot_rhessys(outname = "out/Ntests/HPB_spinup350yr_run4",   date_filter = ymd("2000-01-01"),
                  vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3",
                           "gwNO3", "gwNO3out", "surface_NO3") ) 



# read_plot_rhessys("out/Spin/Spin_valid2/HPB_spinup450yr", validate = TRUE)


```




## Look at coweeta output

```{r}
coweetapath <- "C:/Users/dwh18/Downloads/416d1d3b-7e69-4c4b-ac7f-1d06bb8b6944/rhessysWS18/rhessys_ws18_local/output/rhessysLOCAL"


zzz <- readin_rhessys_output(pre = coweetapath, read_grow = T)

coweetapathgrow <- "C:/Users/dwh18/Downloads/416d1d3b-7e69-4c4b-ac7f-1d06bb8b6944/rhessysWS18/rhessys_ws18_local/output/rhessysLOCAL"

bdg <- read.table(grow_file, header = TRUE)


```






## Run for a few years off the tests

```{r}
# #Can run all of this within the Docker container after opening R
# 
# setwd("repo/HPB_files_NEW")
# library(RHESSysIOinR)
# 
# dates = c("2015 1 1 1", "2026 1 2 24") #has to end with a 24 hour so that the tec output will work
# 
# name = "hpb_spinup_Hydro"
# 
# input_rhessys = IOin_rhessys_input(
#   version          = "../../rhessys7.5",
#   tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
#   world_file       = "worldfiles/NO3GWtests/hpb.world4.Y2300M1D2H23.state", #what was made in pre-processing
#   world_hdr_prefix = "hpb_spintest",
#   flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
#   start            = dates[1],
#   end              = dates[2],
#   output_folder    = "out/NO3_GW_tests/ValidRun4",
#   output_prefix    = "HPB_spinup250yr",
#   commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
# )
# 
# #header directly copied from hpb.hdr
# input_hdr = IOin_hdr(
#   basin      = "defs/basin.def",
#   hillslope  = "defs/hill.def",
#   zone       = "defs/zone.def",
#   soil       = "defs/soil_siltyloam.def",
#   landuse    = "defs/lu_undev.def",
#   stratum    = "defs/veg_deciduous.def",
#   basestations = "clim/ccr_base"
# )
# 
# #set param list
# pars_list = list(
#   list(input_hdr$soil_def[1], "m",                  c(0.108)),
#   list(input_hdr$soil_def[1], "Ksat_0",             c(404.3)),
#   list(input_hdr$soil_def[1], "sat_to_gw_coeff",    c(0)), #c(0.1, 0.1, 0, 0)
#   list(input_hdr$hillslope_def[1], "gw_loss_coeff", c(0)), #c(0.9, 0.9, 0, 0)
#   list(input_hdr$soil_def[1], "NO3_adsorption_rate", c(0.9)) #c(0.1, 0.9, 0.1, 0.9)    #0 is default
#     )
# 
# # #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
# input_def_pars = IOin_def_pars_simple(pars_list)
# 
# 
# #make tec file locally
# input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = F)
# 
# 
# #run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
#   def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output
#   return_cmd    = F   # flip to FALSE to run
# )


```


```{r}
read_plot_rhessys("out/NO3_GW_tests/ValidRun1/HPB_spinup250yr", validate = TRUE)


```








