---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---


docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop



## run using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2500 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup"

## possible way to pull best world file
# Use your best current .state file as starting point
# worldfile_start = tail(list.files("worldfiles", pattern = "\\.state", full.names = TRUE), 1)
# message("Starting from: ", worldfile_start) ## would then set world)file in input rhessys to worldfile_start

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/",
  output_prefix    = "HPB_spinup550yr_multiT3",
  commandline_options = "-b -g -climrepeat" 
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#p0 = c("epc.root_distrib_parm", "epc.min_percent_leafg", "epc.branch_turnover", "epc.froot_turnover", 
#       "epc.leaf_turnover", "epc.livewood_turnover", "epc.gl_c", "epc.gl_smax", "epc.proj_sla")

#where you can manually tweak params
# pars_list = list(
#   # list(def_file_path,       "parameter_name",    value) 
#   list(input_hdr$stratum_def[1], "epc.root_distrib_parm", 18), #default is 8 in deciduous
#   list(input_hdr$stratum_def[1], "epc.leaf_turnover", 1), #default is 1 in deciduous; range is 0-1: %  leaves that die each year
#   list(input_hdr$stratum_def[1], "epc.livewood_turnover", 0.7) #default is 0.7 in deciduous
#   # add as many as you need in the same format
# )


# pars_list = list(list(input_hdr$stratum_def[1], "epc.root_distrib_parm",    c(1, 8, 20)),
#   list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.1, 0.5, 1)))

pars_list = list(
  
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm",    c(20, 20, 20)),
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(1, 1, 1)),
  
  
  # --- Reduce plant C accumulation ---
  
  # Increase wood turnover to remove structural C faster (default 0.7, try higher)
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",   c(0.7, 0.85, 1.0)),
  
  # Turn on branch turnover (default is 0.0 = off)
  list(input_hdr$stratum_def[1], "epc.branch_turnover",     c(0.0, 0.1, 0.25)),
  
  # # Reduce stem allocation relative to leaves (default 1.62)
  # # lower = less structural C accumulation
  # list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",   c(0.8, 1.2, 1.62)),
  # 
  # # Cap NSC storage pool (default 0.2) — lower = less C hoarded in plant
  # list(input_hdr$stratum_def[1], "epc.max_storage_percent",  c(0.1, 0.15, 0.2)),
  
  # --- Increase soil C inputs ---
  
  # Fine root turnover is major soil C pathway (default 0.27, try higher)
  list(input_hdr$stratum_def[1], "epc.froot_turnover",      c(0.27, 0.5, 0.8))
  
  # # Leaf turnover feeds litter → soil (you found this didn't matter much,
  # # but worth including at higher end given soil C deficit)
  # list(input_hdr$stratum_def[1], "epc.leaf_turnover",       c(0.5, 0.75, 1.0))
)

input_def_pars = IOin_def_pars_simple(pars_list)

#set up manually tweak params for rhessys format; can also put multi here
input_def_pars = IOin_def_pars_simple(pars_list)
#when you run this it will geneate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change orignial files


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)



#run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #set in input rhessys; can be NULL if calling a local file and not using IOin_tec_std
#   def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output 
#   return_cmd    = F   # flip to FALSE to run
# )


#run for multi param
run_rhessys_multi(
  input_rhessys   = input_rhessys,
  hdr_files       = input_hdr,
  tec_data        = input_tec_data,
  def_pars        = input_def_pars,
  output_filter   = NULL,
  all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
  return_cmd      = F
)

##seems like it runs by pair for each year: ie 57,57,57, 58,58,58 . at some point it seems to get off tho.. ie 2272, 2259, 2237
## started ~632 - 657




```



## look at output


```{r}
##functin to read in and plot quickly

library(RHESSysIOinR)
library(tidyverse)


read_plot_rhessys = function(outname, date_filter = NULL, ncol = 3,
                              vars = c("lai.x", "plantc.x", "soilc", "plantn",
                                       "streamflow", "unsat_stor",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
  
  dat = readin_rhessys_output(outname)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  if (!is.null(date_filter)) out = out |> filter(date > date_filter)
  
  out |>
    select(date, year, yd, any_of(vars)) |>
    pivot_longer(-c(date, year, yd)) |>
    ggplot(aes(x = date, y = value)) +
    geom_point() +
    facet_wrap(~name, scales = "free_y", ncol = ncol) +
    labs(title = outname, x = "Date", y = NULL) +
    theme_bw()
}

# read_plot_rhessys(outname = "out/Trial1/HPB_spinup350yr_multi_run1", date_filter = NULL)


```




## Latin cube trial 

LHS with RHESSys multi

```{r}

library(lhs)          # install.packages("lhs")
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")


# -------------------- Define parameter ranges --------------------
param_ranges = list(
  epc.root_distrib_parm = c(1,     20),  #8 
  epc.alloc_stemc_leafc = c(1,   5),  # 2.2
  epc.livewood_turnover = c(0.15,   0.9), # 0.7
  epc.froot_turnover    = c(0.4,   1),  # 1
  epc.branch_turnover   = c(0.01,  0.2),  #0
  DOM_decay_rate        = c(0.01, 0.1)  #0.05
)

n_params = length(param_ranges)
n_runs   = 24  #have 14 cores to work with; have been running in batches of 12 for this

# -------------------- Generate LHS design --------------------
set.seed(42)
lhs_raw    = randomLHS(n_runs, n_params)
lhs_scaled = as.data.frame(lhs_raw)
colnames(lhs_scaled) = names(param_ranges)

for (p in names(param_ranges)) {
  lo = param_ranges[[p]][1]
  hi = param_ranges[[p]][2]
  lhs_scaled[[p]] = lo + lhs_scaled[[p]] * (hi - lo)
}

print(head(lhs_scaled))
print(summary(lhs_scaled))

# -------------------- Setup --------------------

dates          = c("1950 1 1 1", "2200 1 2 24")
name           = "hpb_LHS"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = paste0("tecfiles/", name, ".tec"),
  world_file       = "worldfiles/hpb.world",
  world_hdr_prefix = name,
  flowtable        = "flowtables/hpb.flow",
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/LHS_T1",
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat"
)

input_hdr = IOin_hdr(
  basin        = "defs/basin.def",
  hillslope    = "defs/hill.def",
  zone         = "defs/zone.def",
  soil         = "defs/soil_siltyloam.def",
  landuse      = "defs/lu_undev.def",
  stratum      = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], 
                               output_state = FALSE)

# -------------------- Pass full vectors to run_rhessys_multi --------------------
# No [i] indexing — pass the entire column as a vector
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  ## sampled LHS veg and soil params
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", lhs_scaled$epc.root_distrib_parm),
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",  lhs_scaled$epc.alloc_stemc_leafc),
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",  lhs_scaled$epc.livewood_turnover),
  list(input_hdr$stratum_def[1], "epc.froot_turnover",     lhs_scaled$epc.froot_turnover),
  list(input_hdr$stratum_def[1], "epc.branch_turnover",    lhs_scaled$epc.branch_turnover),
  list(input_hdr$soil_def[1],    "DOM_decay_rate",          lhs_scaled$DOM_decay_rate)
)

input_def_pars = IOin_def_pars_simple(pars_list)

# -------------------- Run --------------------
run_rhessys_multi(
  input_rhessys    = input_rhessys,
  hdr_files        = input_hdr,
  tec_data         = input_tec_data,
  def_pars         = input_def_pars,
  output_filter    = NULL,
  all_combinations = FALSE,
  return_cmd       = F   # flip to FALSE to run
)

# -------------------- Save param table --------------------
lhs_scaled |>
  mutate(run = paste0(name, "_run", seq_len(n_runs))) |>
  write_csv("out/LHS_T1/lhs_parameter_table.csv")



```




evaluate LHS

```{r}
library(tidyverse)
library(data.table)
library(RHESSysIOinR)



# Find only the non-grow basin daily files to get base prefixes

#### UPDATE FILE PATH
out_files = list.files("out/LHS_trial5", pattern = "hpb_lhs_run.*basin\\.daily$", 
                        full.names = TRUE)
out_files = out_files[!grepl("grow", out_files)]  # exclude the grow files

all_ts = map_dfr(out_files, function(f) {
  
  # Base prefix is everything before _basin.daily
  # e.g. out/hpb_lhs_run001_basin.daily -> out/hpb_lhs_run001
  run_prefix = gsub("_basin\\.daily$", "", f)
  run        = gsub(".*/", "", run_prefix)
  
  message("Reading: ", run)
  
  dat = tryCatch(
    readin_rhessys_output(run_prefix),
    error = function(e) { 
      message("  Failed: ", e$message)
      return(NULL) 
    }
  )
  
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  out |>
    select(date, year, month, day,
           plantc = plantc.x,
           soilc,
           plantn,
           soiln,
           streamflow_DOC,
           streamflow_DON,
           streamflow_NO3,
           lai    = lai.x,
           unsat_stor, 
           sat_def,
           streamflow) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    mutate(run = run)
})


# -------------------- Read data --------------------
#### UPDATE FILE PATH
param_table = read_csv("out/LHS_trial5/lhs_parameter_table.csv")


# -------------------- Scoring table (unchanged) --------------------
results = all_ts |>
  group_by(run) |>
  slice_tail(n = 365) |>
  summarise(
    mean_plantc     = max(plantc,     na.rm = TRUE),
    mean_soilc      = max(soilc,      na.rm = TRUE),
    mean_lai        = max(lai,         na.rm = TRUE),
    mean_streamflow = max(streamflow,  na.rm = TRUE),
    mean_streamflow_DOC = max(streamflow_DOC,  na.rm = TRUE),
    mean_streamflow_NO3 = max(streamflow_NO3,  na.rm = TRUE)
  ) |>
  left_join(param_table, by = "run") |>
  mutate(score = abs(mean_plantc - 15) + 
                 abs(mean_soilc  - 15)) |> #using 15 as target kg/m2 for both pools curretnly
  arrange(score)

print(head(results, 10))

# -------------------- Time series plot — all runs overlaid --------------------
# Color by score so best runs stand out
all_ts |>
  filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow_DOC, streamflow, soil_moisture),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")

# -------------------- Highlight just the top N runs --------------------
top_n = 5

top_runs = results$run[1:top_n]

all_ts |>
  filter(run %in% top_runs) |>
  left_join(results |> select(run, score), by = "run") |>
  mutate(run_label = paste0(run, "\nscore=", round(score, 2))) |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, color = run_label)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title    = paste("Top", top_n, "LHS runs by score"),
       subtitle = "Are stocks leveling off by end of run?",
       x = "Date", y = NULL, color = "Run") +
  theme_bw() +
  theme(legend.position = "bottom")

# -------------------- Stabilization check --------------------
# Compare mean of first vs last year to see if stocks are still drifting
all_ts |>
  left_join(results |> select(run, score), by = "run") |>
  group_by(run, score) |>
  mutate(run_year = year(as.Date(date)),
         first_year = min(run_year),
         last_year  = max(run_year)) |>
  summarise(
    plantc_start = mean(plantc[run_year == first_year], na.rm = TRUE),
    plantc_end   = mean(plantc[run_year == last_year],  na.rm = TRUE),
    soilc_start  = mean(soilc[run_year  == first_year], na.rm = TRUE),
    soilc_end    = mean(soilc[run_year  == last_year],  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    plantc_drift = round(plantc_end - plantc_start, 3),
    soilc_drift  = round(soilc_end  - soilc_start,  3)
  ) |>
  arrange(score) |>
  select(run, score, plantc_start, plantc_end, plantc_drift, 
         soilc_start, soilc_end, soilc_drift) |>
  print(n = 20)


```


```{r}
all_ts |>
  filter(date > ymd("2200-01-01")) |> 
  filter(streamflow_DOC < 0.1) |> 
  left_join(results |> select(run, score), by = "run") |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow_DOC, streamflow, soil_moisture),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")

```





##### Mega LHS
expanding range of params and values were going across gonna do 48 iterations


```{r}

library(lhs)          # install.packages("lhs")
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")


# -------------------- Define parameter ranges --------------------
param_ranges = list(
  ##hydro (note three are in soil def one in hillsope)
  # m                  = c(0.05,     2),  #0.12
  # Ksat_0            = c(1,   600),  # 0.622
  # sat_to_gw_coeff   = c(0,   1), # 1
  # gw_loss_coeff     = c(0,   1),  # 1; 0 in testing?
  
  ##veg (veg def)
  # P0 from will
  epc.root_distrib_parm = c(6,    20),   # 8 is veg decid default; 19 is github
  epc.min_percent_leafg = c(0,  2),          # def = 1; git says 0.01 to 0.001
  epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1
  epc.froot_turnover    = c(0.6,  0.9),    # def = 1; 0 to 1
  epc.leaf_turnover     = c(0.1,   1.0),  # def = 1; 0 to 1
  epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  epc.gl_c              = c(0.00004, 0.00008),      #def is 0.00006; no range given.         leafscale cuticular conductance
  epc.gl_smax          = c(0.004, 0.009), # def is 0.006;  no range given.         stomatal conductance → more CO2 uptake
  epc.proj_sla         = c(1,    60),    # def is 32 in decid; 9 is git default specific leaf area → more leaf area per C
  #DWH added
  epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more C fixation
  epc.alloc_frootc_leafc = c(0.8, 1.6), # 1.2 and 1.325 by def and git less to fine roots = more structural C
  
  ##soil (soil def)
  DOM_decay_rate        = c(0.01, 0.1),  #0.05
  NO3_adsorption_rate   = c(0, 1),       #0
  N_decay               = c(0.05, 0.3), #0.12
  DON_production_rate   = c(0.1, 0.5)   #0.3
  )

n_params = length(param_ranges)
n_runs   = 48  #have 14 cores to work with; have been running in batches of 12 for this

# -------------------- Generate LHS design --------------------
set.seed(42)
lhs_raw    = randomLHS(n_runs, n_params)
lhs_scaled = as.data.frame(lhs_raw)
colnames(lhs_scaled) = names(param_ranges)

for (p in names(param_ranges)) {
  lo = param_ranges[[p]][1]
  hi = param_ranges[[p]][2]
  lhs_scaled[[p]] = lo + lhs_scaled[[p]] * (hi - lo)
}

print(head(lhs_scaled))
print(summary(lhs_scaled))

# -------------------- Setup --------------------

dates          = c("1950 1 1 1", "2400 1 2 24")
name           = "hpb_LHSmega450"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = paste0("tecfiles/", name, ".tec"),
  world_file       = "worldfiles/hpb.world",
  world_hdr_prefix = name,
  flowtable        = "flowtables/hpb.flow",
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/LHSmega1",
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat"
)

input_hdr = IOin_hdr(
  basin        = "defs/basin.def",
  hillslope    = "defs/hill.def",
  zone         = "defs/zone.def",
  soil         = "defs/soil_siltyloam.def",
  landuse      = "defs/lu_undev.def",
  stratum      = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], 
                               output_state = FALSE)

# -------------------- Pass full vectors to run_rhessys_multi --------------------
# No [i] indexing — pass the entire column as a vector
pars_list = list(
  
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 1),  #0.108
  list(input_hdr$soil_def[1], "Ksat_0",  10),  #404.3
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0), #0.1
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0), #0.9
  # list(input_hdr$soil_def[1], "m", lhs_scaled$m),
  # list(input_hdr$soil_def[1], "Ksat_0",  lhs_scaled$Ksat_0),
  # list(input_hdr$soil_def[1], "sat_to_gw_coeff",  lhs_scaled$sat_to_gw_coeff),
  # list(input_hdr$hillslope_def[1], "gw_loss_coeff", lhs_scaled$gw_loss_coeff),

  ## sampled LHS veg params
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", lhs_scaled$epc.root_distrib_parm),
  list(input_hdr$stratum_def[1], "epc.min_percent_leafg", lhs_scaled$epc.min_percent_leafg),
  list(input_hdr$stratum_def[1], "epc.branch_turnover",    lhs_scaled$epc.branch_turnover),
  list(input_hdr$stratum_def[1], "epc.froot_turnover",     lhs_scaled$epc.froot_turnover),
  list(input_hdr$stratum_def[1], "epc.leaf_turnover",      lhs_scaled$epc.leaf_turnover),
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",  lhs_scaled$epc.livewood_turnover),
  list(input_hdr$stratum_def[1],  "epc.gl_c",                lhs_scaled$epc.gl_c),
  list(input_hdr$stratum_def[1],    "epc.gl_smax",          lhs_scaled$epc.gl_smax),
  list(input_hdr$stratum_def[1],    "epc.proj_sla",          lhs_scaled$epc.proj_sla),
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",   lhs_scaled$epc.alloc_stemc_leafc),
  list(input_hdr$stratum_def[1],  "epc.flnr",              lhs_scaled$epc.flnr),
  list(input_hdr$stratum_def[1], "epc.alloc_frootc_leafc",  lhs_scaled$epc.alloc_frootc_leafc),

  ## sampled LHS veg soil params
  list(input_hdr$soil_def[1],    "DOM_decay_rate",               lhs_scaled$DOM_decay_rate),
  list(input_hdr$soil_def[1],    "NO3_adsorption_rate",          lhs_scaled$NO3_adsorption_rate),
  list(input_hdr$soil_def[1],    "N_decay",                      lhs_scaled$N_decay),
  list(input_hdr$soil_def[1],    "DON_production_rate",          lhs_scaled$DON_production_rate)
)

input_def_pars = IOin_def_pars_simple(pars_list)

# -------------------- Run --------------------
run_rhessys_multi(
  input_rhessys    = input_rhessys,
  hdr_files        = input_hdr,
  tec_data         = input_tec_data,
  def_pars         = input_def_pars,
  output_filter    = NULL,
  all_combinations = FALSE,
  return_cmd       = F   # flip to FALSE to run
)

# -------------------- Save param table --------------------
lhs_scaled |>
  mutate(run = paste0(name, "_run", seq_len(n_runs))) |>
  write_csv("out/LHSmega1/lhs_parameter_table.csv")




```


EVALUATE

```{r}
library(tidyverse)
library(data.table)
library(RHESSysIOinR)



# Find only the non-grow basin daily files to get base prefixes

#### UPDATE FILE PATH
out_files = list.files("out/LHSmega1", pattern = "hpb_LHSmega450_run.*basin\\.daily$", 
                        full.names = TRUE)
out_files = out_files[!grepl("grow", out_files)]  # exclude the grow files

all_ts = map_dfr(out_files, function(f) {
  
  # Base prefix is everything before _basin.daily
  # e.g. out/hpb_lhs_run001_basin.daily -> out/hpb_lhs_run001
  run_prefix = gsub("_basin\\.daily$", "", f)
  run        = gsub(".*/", "", run_prefix)
  
  message("Reading: ", run)
  
  dat = tryCatch(
    readin_rhessys_output(run_prefix),
    error = function(e) { 
      message("  Failed: ", e$message)
      return(NULL) 
    }
  )
  
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  out |>
    select(date, year, month, day,
           plantc = plantc.x,
           soilc,
           plantn,
           soiln,
           streamflow_DOC,
           streamflow_DON,
           streamflow_NO3,
           lai    = lai.x,
           unsat_stor, 
           sat_def,
           streamflow) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    mutate(run = run)
})


# -------------------- Read data --------------------
#### UPDATE FILE PATH
param_table = read_csv("out/LHSmega1/lhs_parameter_table.csv")


# -------------------- Scoring table (unchanged) --------------------
results = all_ts |>
  group_by(run) |>
  slice_tail(n = 365) |>
  summarise(
    mean_plantc     = max(plantc,     na.rm = TRUE),
    mean_soilc      = max(soilc,      na.rm = TRUE),
    mean_lai        = max(lai,         na.rm = TRUE),
    mean_streamflow = max(streamflow,  na.rm = TRUE),
    mean_streamflow_DOC = max(streamflow_DOC,  na.rm = TRUE),
    mean_streamflow_NO3 = max(streamflow_NO3,  na.rm = TRUE)
  ) |>
  left_join(param_table, by = "run") |>
  mutate(score = abs(mean_plantc - 15) + 
                 abs(mean_soilc  - 15)) |> #using 15 as target kg/m2 for both pools curretnly
  arrange(score)

print(head(results, 10))

# -------------------- Time series plot — all runs overlaid --------------------
# Color by score so best runs stand out
all_ts |>
  filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow_DOC, streamflow, soil_moisture),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")

```


```{r}


all_ts |>
  filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
    filter(score > 25, score < 28) |> 
  pivot_longer(cols = c(plantc, soilc, lai, streamflow_DOC, streamflow, soil_moisture),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")




# -------------------- Highlight just the top N runs --------------------
top_n = 5

top_runs = results$run[1:top_n]

all_ts |>
  filter(run %in% top_runs) |>
  filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
  mutate(run_label = paste0(run, "\nscore=", round(score, 2))) |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, color = run_label)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title    = paste("Top", top_n, "LHS runs by score"),
       subtitle = "Are stocks leveling off by end of run?",
       x = "Date", y = NULL, color = "Run") +
  theme_bw() +
  theme(legend.position = "bottom")

# -------------------- Stabilization check --------------------
# Compare mean of first vs last year to see if stocks are still drifting
all_ts |>
  left_join(results |> select(run, score), by = "run") |>
  group_by(run, score) |>
  mutate(run_year = year(as.Date(date)),
         first_year = min(run_year),
         last_year  = max(run_year)) |>
  summarise(
    plantc_start = mean(plantc[run_year == first_year], na.rm = TRUE),
    plantc_end   = mean(plantc[run_year == last_year],  na.rm = TRUE),
    soilc_start  = mean(soilc[run_year  == first_year], na.rm = TRUE),
    soilc_end    = mean(soilc[run_year  == last_year],  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    plantc_drift = round(plantc_end - plantc_start, 3),
    soilc_drift  = round(soilc_end  - soilc_start,  3)
  ) |>
  arrange(score) |>
  select(run, score, plantc_start, plantc_end, plantc_drift, 
         soilc_start, soilc_end, soilc_drift) |>
  print(n = 20)
```









#### SOBOL ####

```{r}

library(sensitivity)   # install.packages("sensitivity")
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")

# -------------------- Parameter ranges --------------------
param_ranges = list(
  m                      = c(0.2, 2),
  Ksat_0                 = c(10,    600),
  sat_to_gw_coeff        = c(0,    1),
  gw_loss_coeff          = c(0,    1),
  epc.root_distrib_parm  = c(6,    20),
  # epc.min_percent_leafg  = c(0,    2),
  # epc.branch_turnover    = c(0.03, 0.04),
  # epc.froot_turnover     = c(0.6,  0.9),
  # epc.leaf_turnover      = c(0.1,  1.0),
  # epc.livewood_turnover  = c(0.15, 0.35), #0.7 by default
  # epc.gl_c               = c(0.00004, 0.00008),
  # epc.gl_smax            = c(0.004,   0.009),
  epc.proj_sla           = c(20,    45),
  # epc.alloc_stemc_leafc  = c(1,    5),
  # epc.flnr               = c(0.05, 0.3),
  # epc.alloc_frootc_leafc = c(0.8,  1.6),
  ## others to look at 3 epc.dead params
  # DOM_decay_rate         = c(0.01, 0.1),
  NO3_adsorption_rate    = c(0,    1)
  # N_decay                = c(0.05, 0.3),
  # DON_production_rate    = c(0.1,  0.5)
)

n_params = length(param_ranges)

# -------------------- Generate Sobol sample matrices --------------------
# n_samples × (n_params + 2) total runs
# 50 samples × 22 = 1100 runs — fast on 1patch, slower on full basin
## DWH trying 7 parma to start and nsamples = 10; those 90 iterations tool ~10 min to run for patches and took up 9GB
n_samples = 50

set.seed(42)
X1 = as.data.frame(matrix(runif(n_samples * n_params), nrow = n_samples))
X2 = as.data.frame(matrix(runif(n_samples * n_params), nrow = n_samples))
colnames(X1) = colnames(X2) = names(param_ranges)

# Scale from [0,1] to actual parameter ranges
for (p in names(param_ranges)) {
  lo = param_ranges[[p]][1]
  hi = param_ranges[[p]][2]
  X1[[p]] = lo + X1[[p]] * (hi - lo)
  X2[[p]] = lo + X2[[p]] * (hi - lo)
}

# Create the Sobol object — this generates the full run matrix
# model = NULL means we run RHESSys manually and feed results back in
sob = sobolSalt(model = NULL, X1 = X1, X2 = X2, 
                scheme = "A", nboot = 100)

# In the container — save sob after creating it
saveRDS(sob, "out/sobol/sob_object.rds")

# sob$X is the full parameter matrix to run — n_samples × (n_params + 2) rows
run_matrix = as.data.frame(sob$X)
colnames(run_matrix) = names(param_ranges)

message("Total runs to execute: ", nrow(run_matrix))
print(head(run_matrix))

# Save run matrix so you can recover it if something crashes
# if (!dir.exists("out/sobol")) dir.create("out/sobol", recursive = TRUE)

write_csv(run_matrix |> mutate(run_id = row_number()),
          "out/sobol/sobol_run_matrix.csv")

# -------------------- Setup RHESSys inputs --------------------

dates = c("1950 1 1 1", "2200 1 2 24")
name  = "hpb_sobol"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = paste0("tecfiles/", name, ".tec"),
  world_file       = "worldfiles/MISC/1patch_spin_KEEP/hpb_1patch_vegID3.world",
  world_hdr_prefix = name,
  flowtable        = "flowtables/1patch_spin/hpb_1patch_vegID3.flow",
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/sobol/",
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat"
)

input_hdr = IOin_hdr(
  basin        = "defs/basin.def",
  hillslope    = "defs/hill.def",
  zone         = "defs/zone.def",
  soil         = "defs/soil_siltyloam.def",
  landuse      = "defs/lu_undev.def",
  stratum      = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

input_tec_data = IOin_tec_std(start = dates[1], end = dates[2],
                               output_state = FALSE)

# -------------------- Build pars_list from run matrix --------------------
# Each column of run_matrix becomes a vector passed to run_rhessys_multi
pars_list = list(
  ### Soil params
  list(input_hdr$soil_def[1],      "m",                   run_matrix$m),
  list(input_hdr$soil_def[1],      "Ksat_0",              run_matrix$Ksat_0),
  list(input_hdr$soil_def[1],      "sat_to_gw_coeff",     run_matrix$sat_to_gw_coeff),
  # list(input_hdr$soil_def[1],      "DOM_decay_rate",      run_matrix$DOM_decay_rate),
  list(input_hdr$soil_def[1],      "NO3_adsorption_rate", run_matrix$NO3_adsorption_rate),
  # list(input_hdr$soil_def[1],      "N_decay",             run_matrix$N_decay),
  # list(input_hdr$soil_def[1],      "DON_production_rate", run_matrix$DON_production_rate),
  ### Hillslope params
  list(input_hdr$hillslope_def[1], "gw_loss_coeff",       run_matrix$gw_loss_coeff),
  ### Stratum params
  list(input_hdr$stratum_def[1],   "epc.root_distrib_parm",  run_matrix$epc.root_distrib_parm),
  # list(input_hdr$stratum_def[1],   "epc.min_percent_leafg",  run_matrix$epc.min_percent_leafg),
  # list(input_hdr$stratum_def[1],   "epc.branch_turnover",    run_matrix$epc.branch_turnover),
  # list(input_hdr$stratum_def[1],   "epc.froot_turnover",     run_matrix$epc.froot_turnover),
  # list(input_hdr$stratum_def[1],   "epc.leaf_turnover",      run_matrix$epc.leaf_turnover),
  # list(input_hdr$stratum_def[1],   "epc.livewood_turnover",  run_matrix$epc.livewood_turnover),
  # list(input_hdr$stratum_def[1],   "epc.gl_c",               run_matrix$epc.gl_c),
  # list(input_hdr$stratum_def[1],   "epc.gl_smax",            run_matrix$epc.gl_smax),
  list(input_hdr$stratum_def[1],   "epc.proj_sla",           run_matrix$epc.proj_sla)
  # list(input_hdr$stratum_def[1],   "epc.alloc_stemc_leafc",  run_matrix$epc.alloc_stemc_leafc),
  # list(input_hdr$stratum_def[1],   "epc.flnr",               run_matrix$epc.flnr),
  # list(input_hdr$stratum_def[1],   "epc.alloc_frootc_leafc", run_matrix$epc.alloc_frootc_leafc)
)

input_def_pars = IOin_def_pars_simple(pars_list)

# -------------------- Run --------------------
run_rhessys_multi(
  input_rhessys    = input_rhessys,
  hdr_files        = input_hdr,
  tec_data         = input_tec_data,
  def_pars         = input_def_pars,
  output_filter    = NULL,
  all_combinations = FALSE,
  return_cmd       = F   # flip to FALSE to run
)


```


EVALUATE

```{r}
source("../Unit_Conversions.R")

#readin cretated run matrix and sob file
run_matrix <- read.csv("D:/RHESSys/OUT_Sobol1/sobol/sobol_run_matrix.csv")
sob = readRDS("D:/RHESSys/OUT_Sobol1/sobol/sob_object.rds")

# -------------------- Read outputs and compute model responses --------------------
# out_files = list.files("out/sobol", pattern = "basin\\.daily$", full.names = TRUE)
out_files = list.files("D:/RHESSys/OUT_Sobol1/sobol", pattern = "basin\\.daily$", full.names = TRUE)
out_files = out_files[!grepl("grow", out_files)]
# out_files = sort(out_files)  # critical — must match row order of run_matrix

message("Files found: ", length(out_files))
message("Runs executed: ", nrow(run_matrix))
# These must be equal before proceeding

all_ts = map_dfr(seq_along(out_files), function(i) {
  f          = out_files[i]
  run_prefix = gsub("_basin\\.daily$", "", f)
  
  dat = tryCatch(readin_rhessys_output(run_prefix),
                 error = function(e) { message("Failed: ", e); return(NULL) })
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  out |>
    mutate(
      streamflow_Ls = mm_to_Ls(streamflow, 4.299446),
      DOC_mgL       = gm2day_to_mgL(streamflow_DOC, streamflow),
      NO3_mgL       = gm2day_to_mgL(streamflow_NO3, streamflow)
    ) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    select(date, lai = lai.x, plantc = plantc.x, soilc, plantn, soiln, soil_moisture,
           streamflow, streamflow_DOC, streamflow_NO3,
           streamflow_Ls, DOC_mgL, NO3_mgL) |>
    mutate(run_id = i)
})



# plot it all
all_ts |>
  mutate(year = year(date)) |> 
  group_by(year, run_id) |> 
  summarize(across(c(lai:NO3_mgL), mean, na.rm = TRUE)) |> 
  select(year, run_id, everything()) |> 
    # filter(date > ymd("2000-01-01")) |> 
  pivot_longer(-c(1:2),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = year, y = value, 
             group = run_id, color = run_id)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "runID") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "sobol patch runs — full time series",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")



# -------------------- Compute one scalar response per run --------------------
# Sobol needs a single number per run — compute several and run Sobol on each
responses = all_ts |>
  group_by(run_id) |>
  summarise(
    # End of run C stocks
    mean_plantc     = mean(tail(plantc,    365), na.rm = TRUE),
    mean_soilc      = mean(tail(soilc,     365), na.rm = TRUE),
    mean_lai        = mean(tail(lai,        365), na.rm = TRUE),
    mean_streamflow = mean(tail(streamflow_Ls, 365), na.rm = TRUE),
    mean_NO3        = mean(tail(NO3_mgL,   365), na.rm = TRUE),
    mean_DOC        = mean(tail(DOC_mgL,   365), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(run_id)   # must match run_matrix row order

# Confirm alignment
stopifnot(nrow(responses) == nrow(run_matrix))

# -------------------- Feed results back to Sobol and get indices --------------------
# Run Sobol separately for each response variable
sobol_results = list()

for (resp_var in c("mean_plantc", "mean_soilc", "mean_lai", 
                    "mean_streamflow", "mean_NO3", "mean_DOC")) {
  
  message("Computing Sobol indices for: ", resp_var)
  
  # Feed model output back into the Sobol object
  sob_copy = sob
  tell(sob_copy, responses[[resp_var]])
  
  # Extract first-order (S1) and total-effect (ST) indices
  sobol_results[[resp_var]] = tibble(
    parameter = names(param_ranges),
    S1        = sob_copy$S$original,    # first order sensitivity
    S1_conf   = sob_copy$S$`max. c.i.` - sob_copy$S$`min. c.i.`,  # confidence interval
    ST        = sob_copy$T$original,    # total effect
    ST_conf   = sob_copy$T$`max. c.i.` - sob_copy$T$`min. c.i.`,
    response  = resp_var
  )
}

sobol_df = bind_rows(sobol_results)

# -------------------- Plot Sobol indices --------------------
# Plot 1: ST indices per response — which params matter for each output?
sobol_df |>
  ggplot(aes(x = reorder(parameter, ST), y = ST, fill = ST)) +
  geom_col() +
  geom_errorbar(aes(ymin = ST - ST_conf/2, ymax = ST + ST_conf/2), 
                width = 0.3) +
  scale_fill_viridis_c(direction = -1) +
  coord_flip() +
  facet_wrap(~response, scales = "free_x", ncol = 3) +
  labs(title = "Sobol total-effect indices (ST) by response variable",
       subtitle = "Higher ST = parameter explains more output variance",
       x = NULL, y = "ST index") +
  theme_bw() +
  theme(legend.position = "none")

# Plot 2: S1 vs ST — reveals parameter interactions
sobol_df |>
  filter(response == "mean_plantc") |>   # change to any response
  ggplot(aes(x = S1, y = ST, label = parameter)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(size = 3) +
  labs(title    = "S1 vs ST — parameter interactions",
       subtitle  = "Points above dashed line interact with other parameters",
       x = "First-order index (S1)",
       y = "Total-effect index (ST)") +
  theme_bw()

# -------------------- Summarize which parameters to keep calibrating --------------------
# Parameters with ST > 0.05 across any response are worth calibrating
important_params = sobol_df |>
  group_by(parameter) |>
  summarise(max_ST = max(ST, na.rm = TRUE),
            responses_sensitive = sum(ST > 0.05, na.rm = TRUE)) |>
  arrange(desc(max_ST))

print(important_params)

message("\nParameters to keep in calibration (ST > 0.05 for any response):")
print(important_params |> filter(max_ST > 0.05) |> pull(parameter))

message("\nParameters that can be fixed at defaults (ST < 0.05 everywhere):")
print(important_params |> filter(max_ST < 0.05) |> pull(parameter))

```




try monthly files for size

```{r}
# out_files = list.files("out/sobol", pattern = "basin\\.daily$", full.names = TRUE)
out_files2 = list.files("D:/RHESSys/OUT_Sobol2/sobol", pattern = "basin\\.monthly$", full.names = TRUE)
out_files2 = out_files2[!grepl("grow", out_files2)]
# out_files = sort(out_files)  # critical — must match row order of run_matrix

message("Files found: ", length(out_files))
message("Runs executed: ", nrow(run_matrix))
# These must be equal before proceeding

all_ts2 = map_dfr(seq_along(out_files2), function(i) {
  f          = out_files2[i]
  run_prefix = gsub("_basin\\.monthly$", "", f)
  
  dat = tryCatch(readin_rhessys_output(run_prefix),
                 error = function(e) { message("Failed: ", e); return(NULL) })
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  out |>
    mutate(
      streamflow_Ls = mm_to_Ls(streamflow, 4.299446),
      DOC_mgL       = gm2day_to_mgL(streamflow_DOC, streamflow),
      NO3_mgL       = gm2day_to_mgL(streamflow_NO3, streamflow)
    ) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    select(date, lai = lai.x, plantc = plantc.x, soilc, plantn, soiln, soil_moisture,
           streamflow, streamflow_DOC, streamflow_NO3,
           streamflow_Ls, DOC_mgL, NO3_mgL) |>
    mutate(run_id = i)
})


# plot it all
all_ts2 |>
  select(date, run_id, everything()) |> 
  filter(date > ymd("2000-01-01")) |> 
  pivot_longer(-c(1:2),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run_id, color = run_id)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "runID") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "sobol patch runs — full time series",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")


```






















