---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---

## set up docker container: run this in the powershell
docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop


## Function to look at outputs

```{r}

#read in validation csv
target_df <- read_csv("../Target_Data_comp/TargetData_15mar26.csv") |> 
  rename(date = Date)

##function to read in and plot quickly
#libraries
library(RHESSysIOinR)
library(tidyverse)
#unit conversion functions
source("../Unit_Conversions.R")

#make function
read_plot_rhessys = function(outname, date_filter = NULL, ncol = 3,
                              vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
  
  dat = readin_rhessys_output(outname)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  if (!is.null(date_filter)) out = out |> filter(date > date_filter)
  
  out |>
    select(date, year, yd, any_of(vars)) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    mutate(streamflow_L_s = mm_to_Ls(streamflow, 4.299446), #HPB area is 4.299446 km^2
           DOC_mgL = gm2day_to_mgL(streamflow_DOC, streamflow),
           NO3_mgL = gm2day_to_mgL(streamflow_NO3, streamflow)) |> 
    pivot_longer(-c(date, year, yd)) |>
    ggplot(aes(x = date, y = value)) +
    geom_point() +
    facet_wrap(~name, scales = "free_y", ncol = ncol) +
    labs(title = outname, x = "Date", y = NULL) +
    theme_bw()
}

# read_plot_rhessys(outname = "out/Trial1/HPB_spinup350yr_multi_run1", date_filter = NULL)

# #make function
# read_plot_validate_rhessys = function(outname, target_data, date_filter = NULL, ncol = 3,
#                               vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
#                                        "streamflow", "unsat_stor", "sat_def",
#                                        "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
#   
#   dat = readin_rhessys_output(outname)
#   
#   out = left_join(dat$bd, dat$bdg,
#                   by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
#   
#   
#   out_target <- left_join(out, target_data, by = c("date"))
#   
# 
#   out_target |>
#     select(date, year, yd, any_of(vars)) |>
#     mutate(soil_moisture =  sat_def - unsat_stor) |> 
#     mutate(streamflow_L_s = mm_to_Ls(streamflow, 4.299446), #HPB area is 4.299446 km^2
#            DOC_mgL = gm2day_to_mgL(streamflow_DOC, streamflow),
#            NO3_mgL = gm2day_to_mgL(streamflow_NO3, streamflow)) |> 
#     pivot_longer(-c(date, year, yd)) |>
#     ggplot(aes(x = date, y = value)) +
#     geom_point() +
#     facet_wrap(~name, scales = "free_y", ncol = ncol) +
#     labs(title = outname, x = "Date", y = NULL) +
#     theme_bw()
# }

```


new function with validation option

```{r}
read_plot_rhessys = function(outname, date_filter = NULL, ncol = 3, validate = FALSE,
                              vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
  
  dat = readin_rhessys_output(outname)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  if (!is.null(date_filter)) out = out |> filter(date > date_filter)
  
  # Base model output with unit conversions
  out_processed = out |>
    select(date, year, yd, any_of(vars)) |>
    mutate(
      soil_moisture    = sat_def - unsat_stor,
      streamflow_L_s   = mm_to_Ls(streamflow, 4.299446),
      DOC_mgL          = gm2day_to_mgL(streamflow_DOC, streamflow),
      NO3_mgL          = gm2day_to_mgL(streamflow_NO3, streamflow)
    )
  
  # -------------------- Standard plot (validate = FALSE) --------------------
  if (!validate) {
    out_processed |>
      pivot_longer(-c(date, year, yd)) |>
      ggplot(aes(x = date, y = value)) +
      geom_point(size = 0.5) +
      facet_wrap(~name, scales = "free_y", ncol = ncol) +
      labs(title = outname, x = "Date", y = NULL) +
      theme_bw()
    
  # -------------------- Validation plot (validate = TRUE) --------------------
  } else {
    
    # if (!file.exists("target_data.csv")) {
    #   stop("target_data.csv not found in working directory: ", getwd())
    # }
    # 
    # target = read_csv("target_data.csv", show_col_types = FALSE) |>
    #   mutate(date = as.Date(date))
    
    target <- target_df
    
    val_start = as.Date("2020-01-01")
    val_end   = as.Date("2026-12-31")
    
    # -------------------- Plot 1: Multi-variable validation --------------------
    # Uses Q_L_s_Flowmate as the streamflow target
    modeled_long = out_processed |>
      filter(date >= val_start, date <= val_end) |>
      select(date,
             lai     = lai.x,
             DOC_mgL,
             NO3_mgL,
             Q_L_s   = streamflow_L_s) |>
      pivot_longer(-date, names_to = "variable", values_to = "modeled")
    
    target_long = target |>
      filter(date >= val_start, date <= val_end) |>
      select(date,
             lai     = lai_MODIS,
             DOC_mgL,
             NO3_mgL,
             Q_L_s   = Q_L_s_Flowmate) |>   # Flowmate for main validation plot
      pivot_longer(-date, names_to = "variable", values_to = "observed")
    
    combined = left_join(modeled_long, target_long, by = c("date", "variable"))
    
    p1 = ggplot(combined, aes(x = date)) +
      geom_line(aes(y = modeled, color = "Modeled"), linewidth = 0.5) +
      geom_point(aes(y = observed, color = "Observed"), size = 1.5, na.rm = TRUE) +
      scale_color_manual(values = c("Modeled"  = "steelblue",
                                     "Observed" = "firebrick")) +
      facet_wrap(~variable, scales = "free_y", ncol = 2) +
      labs(title    = paste("Validation:", outname),
           subtitle = paste("Flowmate Q |", val_start, "to", val_end),
           x = "Date", y = NULL, color = NULL) +
      theme_bw() +
      theme(legend.position = "bottom")
    
    # -------------------- Plot 2: PT streamflow only from 2024 --------------------
    pt_start = as.Date("2024-01-01")
    
    modeled_q = out_processed |>
      filter(date >= pt_start, date <= val_end) |>
      select(date, Q_L_s = streamflow_L_s)
    
    target_q = target |>
      filter(date >= pt_start, date <= val_end) |>
      select(date, Q_L_s_PT)
    
    combined_q = left_join(modeled_q, target_q, by = "date")
    
    p2 = ggplot(combined_q, aes(x = date)) +
      geom_line(aes(y = Q_L_s, color = "Modeled"), linewidth = 0.5) +
      geom_line(aes(y = Q_L_s_PT, color = "Observed (PT)"), 
                 size = 0.5, na.rm = TRUE) +
      scale_color_manual(values = c("Modeled"       = "steelblue",
                                     "Observed (PT)" = "firebrick")) +
      labs(title    = paste("Streamflow validation:", outname),
           subtitle = paste("PT gauge | 2024 onwards"),
           x = "Date", y = "Q (L/s)", color = NULL) +
      theme_bw() +
      theme(legend.position = "bottom")
    
    # Print both plots
    print(p1)
    print(p2)
  }
}

# -------------------- Usage --------------------
# #Standard plot
# read_plot_rhessys("out/Spin/Spin1/HPB_spinup250yr")
# 
# # Validation plot
# read_plot_rhessys("out/Spin/Spin1/HPB_spinup250yr", validate = TRUE)
# 
# # With date filter on the standard plot
# read_plot_rhessys("out/HPB_spinup50yr", date_filter = ymd("2000-01-01"))

```




## run using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2200 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Spin/",
  output_prefix    = "HPB_spinup250yr",
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  #tweak growth params
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)) #1 is default
  
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)


#run for multi param
# run_rhessys_multi(
#   input_rhessys   = input_rhessys,
#   hdr_files       = input_hdr,
#   tec_data        = input_tec_data,
#   def_pars        = input_def_pars,
#   output_filter   = NULL,
#   all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
#   return_cmd      = F
# )


```




## Look at first spin output

```{r}

read_plot_rhessys(outname = "out/Spin/HPB_spinup250yr", 
                  date_filter = ymd("1970-01-01") ) # NULL is default ymd("2000-01-01")


```



## Spin off the spinup

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2200 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/Basin_spin/hpb.world.Y2200M1D2H23.state", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Spin/Spin2/",
  output_prefix    = "HPB_spinup250yr",
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  #tweak growth params
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)) #1 is default
  
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)

```


##look at second spin

```{r}
#Standard plot
read_plot_rhessys("out/Spin/Spin2/HPB_spinup250yr", date_filter = NULL)

# Validation plot
read_plot_rhessys("out/Spin/Spin2/HPB_spinup250yr", validate = TRUE)

```



## Try running for just several years off of each worldfile state

Frist spinup

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("2018 1 1 1", "2026 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/Basin_spin/hpb.world.Y2200M1D2H23.state", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Spin/Spin_valid1/",
  output_prefix    = "HPB_spinup250yr",
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  #tweak growth params
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)) #1 is default
  
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = F)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)

```



```{r}
#Standard plot
read_plot_rhessys("out/Spin/Spin_valid1/HPB_spinup250yr", date_filter = NULL)

# Validation plot
read_plot_rhessys("out/Spin/Spin_valid1/HPB_spinup250yr", validate = TRUE)

```





### LHS based params from L5

```{r}

#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2400 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/hpb.world.Y2400M1D2H23.state", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Spin/Spin4/",
  output_prefix    = "HPB_spinup450yr",
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  #tweak growth params
  #These are *I think* from LHS t5~
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", c(12)),  
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc", c(3)), 
  list(input_hdr$stratum_def[1], "epc.livewood_turnover", c(0.25)),
  list(input_hdr$stratum_def[1], "epc.froot_turnover", c(0.8)),
  list(input_hdr$stratum_def[1], "epc.branch_turnover", c(0.03)),
  list(input_hdr$stratum_def[1], "DOM_decay_rate", c(0.05))
  
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8)), #1 is default
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)


```


```{r}
#Standard plot
read_plot_rhessys("out/Spin/Spin4/HPB_spinup450yr", date_filter = NULL)

# Validation plot
read_plot_rhessys("out/Spin/Spin3/HPB_spinup450yr", validate = TRUE)

```




## Try running for just several years off of each worldfile state

Frist spinup

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("2018 1 1 1", "2026 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  world_file       = "worldfiles/hpb.world.Y2400M1D2H23.state.Y2400M1D2H23.state", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/Spin/Spin_valid2/",
  output_prefix    = "HPB_spinup450yr",
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#set param list
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),
  
  #tweak growth params
  #These are *I think* from LHS t5~
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", c(12)),  
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc", c(3)), 
  list(input_hdr$stratum_def[1], "epc.livewood_turnover", c(0.25)),
  list(input_hdr$stratum_def[1], "epc.froot_turnover", c(0.8)),
  list(input_hdr$stratum_def[1], "epc.branch_turnover", c(0.03)),
  list(input_hdr$stratum_def[1], "DOM_decay_rate", c(0.05))
  # epc.root_distrib_parm = c(8,    12),   # 8 is veg decid default; 19 is github; 8 seemed good
  # epc.proj_sla         = c(25,    40),    # def is 32 in decid; 9 is git default specific leaf area. 
  # some other values to play with 
  # epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1; 0 is in both defs
  # epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  # epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  # epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more 
    )

# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = F)


#run model with inputs defined above
run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files     = input_hdr,
  tec_data      = input_tec_data, #can be NULL if calling a local file and not using IOin_tec_std
  def_pars      = input_def_pars,
  output_filter = NULL, #cant get any version working currently so using legacy output
  return_cmd    = F   # flip to FALSE to run
)

```



```{r}
#Standard plot
read_plot_rhessys("out/Spin/Spin_valid2/HPB_spinup450yr", date_filter = NULL)

# Validation plot
read_plot_rhessys("out/Spin/Spin_valid2/HPB_spinup450yr", validate = TRUE)

```


























