---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---

##Info on command line -s and -gw scalars
```{r}
##info on command line params: https://github.com/RHESSys/RHESSys/wiki/RHESSys-command-line-options
### -s must be followed by two parameter values for m and K
# the first value this is a multiplier on the value set for the m parameter in the soil definition files
# the second required value designates K, hydraulic conductivity at the surface - and is a multiplier on the value set for Ksat_0 in the soil definition files;

### -gw must be followed by two values for gw1 and gw2, respectively. 
#The first value is a multiplier on the sat_to_gw_coeff parameter set in the soil definition file (representing the amount of water moving from the saturated store to the groundwater store). 
#The second value is a multiplier on the gw_loss_coeff parameter in the hillslope default file (representing the amount of water moving from the groundwater store to the stream).

#defaults are: m = 0.12; K_sat0 = 0.622; sat_to_gw_coeff = 1; gw_loss_coeff = 1

```

docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop



## run using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2500 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup"

## possible way to pull best world file
# Use your best current .state file as starting point
# worldfile_start = tail(list.files("worldfiles", pattern = "\\.state", full.names = TRUE), 1)
# message("Starting from: ", worldfile_start) ## would then set world)file in input rhessys to worldfile_start

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  # tec_file         = "tecfiles/tec50yr.test", #call local tecfile
  world_file       = "worldfiles/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/",
  output_prefix    = "HPB_spinup550yr_multiT3",
  commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9 -climrepeat"
  #commandline_options = "-b -g -s 1 100 -gw 1 1"
  # swap to below for neutral spinup params:
  # commandline_options = "-b -g -s 1 10 -gw 0 0"
  # commandline_options = "-b -g"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

#p0 = c("epc.root_distrib_parm", "epc.min_percent_leafg", "epc.branch_turnover", "epc.froot_turnover", 
#       "epc.leaf_turnover", "epc.livewood_turnover", "epc.gl_c", "epc.gl_smax", "epc.proj_sla")

#where you can manually tweak params
# pars_list = list(
#   # list(def_file_path,       "parameter_name",    value) 
#   list(input_hdr$stratum_def[1], "epc.root_distrib_parm", 18), #default is 8 in deciduous
#   list(input_hdr$stratum_def[1], "epc.leaf_turnover", 1), #default is 1 in deciduous; range is 0-1: %  leaves that die each year
#   list(input_hdr$stratum_def[1], "epc.livewood_turnover", 0.7) #default is 0.7 in deciduous
#   # add as many as you need in the same format
# )


# pars_list = list(list(input_hdr$stratum_def[1], "epc.root_distrib_parm",    c(1, 8, 20)),
#   list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.1, 0.5, 1)))

pars_list = list(
  
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm",    c(20, 20, 20)),
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(1, 1, 1)),
  
  
  # --- Reduce plant C accumulation ---
  
  # Increase wood turnover to remove structural C faster (default 0.7, try higher)
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",   c(0.7, 0.85, 1.0)),
  
  # Turn on branch turnover (default is 0.0 = off)
  list(input_hdr$stratum_def[1], "epc.branch_turnover",     c(0.0, 0.1, 0.25)),
  
  # # Reduce stem allocation relative to leaves (default 1.62)
  # # lower = less structural C accumulation
  # list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",   c(0.8, 1.2, 1.62)),
  # 
  # # Cap NSC storage pool (default 0.2) — lower = less C hoarded in plant
  # list(input_hdr$stratum_def[1], "epc.max_storage_percent",  c(0.1, 0.15, 0.2)),
  
  # --- Increase soil C inputs ---
  
  # Fine root turnover is major soil C pathway (default 0.27, try higher)
  list(input_hdr$stratum_def[1], "epc.froot_turnover",      c(0.27, 0.5, 0.8))
  
  # # Leaf turnover feeds litter → soil (you found this didn't matter much,
  # # but worth including at higher end given soil C deficit)
  # list(input_hdr$stratum_def[1], "epc.leaf_turnover",       c(0.5, 0.75, 1.0))
)

input_def_pars = IOin_def_pars_simple(pars_list)

#set up manually tweak params for rhessys format; can also put multi here
input_def_pars = IOin_def_pars_simple(pars_list)
#when you run this it will geneate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change orignial files


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)



#run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #set in input rhessys; can be NULL if calling a local file and not using IOin_tec_std
#   def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output 
#   return_cmd    = F   # flip to FALSE to run
# )


#run for multi param
run_rhessys_multi(
  input_rhessys   = input_rhessys,
  hdr_files       = input_hdr,
  tec_data        = input_tec_data,
  def_pars        = input_def_pars,
  output_filter   = NULL,
  all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
  return_cmd      = F
)

##seems like it runs by pair for each year: ie 57,57,57, 58,58,58 . at some point it seems to get off tho.. ie 2272, 2259, 2237
## started ~632 - 657




```



## look at output


```{r}
##functin to read in and plot quickly

library(RHESSysIOinR)
library(tidyverse)


read_plot_rhessys = function(outname, date_filter = NULL, ncol = 3,
                              vars = c("lai.x", "plantc.x", "soilc", "plantn",
                                       "streamflow", "unsat_stor",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
  
  dat = readin_rhessys_output(outname)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  if (!is.null(date_filter)) out = out |> filter(date > date_filter)
  
  out |>
    select(date, year, yd, any_of(vars)) |>
    pivot_longer(-c(date, year, yd)) |>
    ggplot(aes(x = date, y = value)) +
    geom_point() +
    facet_wrap(~name, scales = "free_y", ncol = ncol) +
    labs(title = outname, x = "Date", y = NULL) +
    theme_bw()
}

# read_plot_rhessys(outname = "out/Trial1/HPB_spinup350yr_multi_run1", date_filter = NULL)


```




##try a spinup off Trial3_run3 that got soil and plant stocks ~=

```{r}

#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2300 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup"


input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  # world_file       = "worldfiles/hpb.world", #what was made in pre-processing
  world_file       = "out/Trial3/hpb.world3.Y2500M1D2H23.state", #what was made in pre-processing #get T3run3 world
  world_hdr_prefix = "hpb_spintestT3r3",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/",
  output_prefix    = "HPB_spinup350yr_offT3r3",
  commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9 -climrepeat"
 )

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)



pars_list = list(
  
  list(input_hdr$stratum_def[1], "epc.root_distrib_parm",    c(5, 10, 20)),
  list(input_hdr$stratum_def[1], "epc.leaf_turnover", c(0.8, 0.9, 1)),
  
  
  # --- Reduce plant C accumulation ---
  
  # Increase wood turnover to remove structural C faster (default 0.7, try higher)
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",   c(0.7, 0.85, 1.0)),
  
  # Turn on branch turnover (default is 0.0 = off)
  list(input_hdr$stratum_def[1], "epc.branch_turnover",     c(0.1, 0.1, 0.1)),
  
  list(input_hdr$stratum_def[1], "epc.proj_sla",           c(9, 15, 25)),
  
  # # Reduce stem allocation relative to leaves (default 1.62)
  # # lower = less structural C accumulation
  # list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",   c(0.8, 1.2, 1.62)),
  # 
  # # Cap NSC storage pool (default 0.2) — lower = less C hoarded in plant
  # list(input_hdr$stratum_def[1], "epc.max_storage_percent",  c(0.1, 0.15, 0.2)),
  
  # --- Increase soil C inputs ---
  
  # Fine root turnover is major soil C pathway (default 0.27, try higher)
  list(input_hdr$stratum_def[1], "epc.froot_turnover",      c(0.4, 0.4, 0.4))
  
  # # Leaf turnover feeds litter → soil (you found this didn't matter much,
  # # but worth including at higher end given soil C deficit)
  # list(input_hdr$stratum_def[1], "epc.leaf_turnover",       c(0.5, 0.75, 1.0))
)

input_def_pars = IOin_def_pars_simple(pars_list)

#set up manually tweak params for rhessys format; can also put multi here
input_def_pars = IOin_def_pars_simple(pars_list)
#when you run this it will geneate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change orignial files


#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)



#run for multi param
run_rhessys_multi(
  input_rhessys   = input_rhessys,
  hdr_files       = input_hdr,
  tec_data        = input_tec_data,
  def_pars        = input_def_pars,
  output_filter   = NULL,
  all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
  return_cmd      = F
)




```



```{r}

read_plot_rhessys(outname = "out/Trial3r3_respin/HPB_spinup350yr_offT3r3_run1", date_filter = NULL ) # ymd("2000-01-01")

read_plot_rhessys(outname = "out/Trial3r3_respin/HPB_spinup350yr_offT3r3_run2", date_filter = NULL)

read_plot_rhessys(outname = "out/Trial3r3_respin/HPB_spinup350yr_offT3r3_run3", date_filter = NULL)




```





----------- output filter trials that aren't working -----------------

```{r}
##output filter
# source("https://raw.githubusercontent.com/wburke24/rhutils/refs/heads/main/R/output_aliases.R")

# output_vars_minimal = c("patch.soil_cs.totalc", "patch.soil_ns.totaln", "stratum.cs.net_psn", 
#                         "patch.sat_deficit", "patch.lai", "patch.totalc", "patch.evaporation", "patch.streamflow")
# 
# outfilter = build_output_filter(
#   timestep = "daily",
#   output_format = "csv",
#   output_path = "output",
#   output_filename = paste0(name,"_basin"),
#   spatial_level = "basin",
#   spatial_ID = 1,
#   variables = output_vars_minimal
# )
# output_filter = IOin_output_filters(outfilter, file_name = "./out/filters/filter2.yml")

# outfilter = build_output_filter(
#   timestep = "daily",
#   output_format = "csv",
#   output_path = "output",
#   output_filename = name,
#   spatial_level = "basin",
#   spatial_ID = 1,
#   variables = c("patch.soil_cs.totalc", "patch.soil_ns.totaln", "patch.lai", "stratum.cs.totalc", "stratum.cs.leafc", "stratum.cs.live_stemc")
# )
# output_filter = IOin_output_filters(outfilter, file_name = "./out/filters/filter.yml")

# if (!dir.exists("out/filters")) dir.create("out/filters", recursive = TRUE)
# 
# yaml_content = 'filter:
#   timestep: daily
#   output:
#     format: csv
#     path: "out/"
#     filename: "hpb_spinup"
#   basin:
#     ids: 1
#     variables: [patch.soil_cs.totalc, patch.soil_ns.totaln, patch.lai, stratum.cs.totalc, stratum.cs.leafc, stratum.cs.live_stemc]'
# 
# writeLines(yaml_content, "out/filters/filter.yml")
# 
# # Verify it looks correct before running
# readLines("out/filters/filter.yml")

```




--------------------- OLD ways of running that work --------------------------------------------


## Running using run_rhessys_single with kmeans patches
this is for HPB_files_NEW folder that uses the kmeans500 patches 
ran without resprouts after 1955 but had huge veg stocks

```{r}

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = "tecfiles/tec.test",
  world_file       = "worldfiles/hpb.world",
  world_hdr_prefix = "hpb",
  flowtable        = "flowtables/hpb.flow",
  start            = "1950 01 1 1",
  end              = "2025 06 1 1",
  output_folder    = "out/",
  output_prefix    = "HPB_spinup75yr",
  commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9"
  # swap to below for neutral spinup params:
  # commandline_options = "-b -g -s 1 10 -gw 0 0"
)

run_rhessys_single(
  input_rhessys = input_rhessys,
  hdr_files = "worldfiles/hpb.hdr",
  # tec_data  = NULL,   def_pars = NULL,  output_filter = NULL,
  return_cmd    = F   # flip to FALSE to run; T prints the command line structure
)

## to run using this world file output as a spinup, make sure tec files specifies the output; and then run again but update world file to be something like "worldfiles/hpb.world.Y2024M12D31.state"


```



## Running old style command line with kmeans patches
ALL OF THIS HAPPENS IN THE DOCKER CONTAINER

```{r}

# Spinup params per wiki: m=1, K=10, gw1=0, gw2=0
# -s m K   → -s 1 10
# -gw gw1 gw2 → -gw 0 0
# 75 years: 1950 to 2025
# -g = grow mode (required for spinup so veg/soil C & N can evolve)
# NO -b flag during spinup — output_current_state handles the worldfile output

## Prior cmd2: cmd2 = sprintf("-pre out/HPBtest_kmean -s 0.9 650, -gw 0.1 0.9")

## adding : 2025 05 1 24 output_current_state
#to the tec file outputs the worldfile

# In the R session with the container set working directory
setwd("repo/HPB_files_NEW")

## These lines set up where to find RHESSys model version and config files 
rhessys_ver = "../../rhessys7.5" #to point backwards to model
worldfile = "worldfiles/hpb.world"
header = "worldfiles/hpb.hdr"
flowtable = "flowtables/hpb.flow"
tecfile = "tecfiles/tec.test" 

## These set up the command line to tell the model what time frame to run and what outputs to set 
#If you want to change duration you also have to update the tec.test file to match cmd3
cmd1 = sprintf("%s -t %s -w %s -whdr %s -r %s",
                     rhessys_ver, tecfile, worldfile, header, flowtable)
cmd2 = sprintf("-pre out/HPB_spinup75yr -s 0.9 650, -gw 0.1 0.9")
# cmd2 = sprintf("-pre out/HPB_spinup75yr -s 1 10 -gw 0 0")
cmd3 = sprintf("-st 1950 01 1 1 -ed 2025 06 1 1 -b -g")
command_line = paste(cmd1,cmd2,cmd3)

#This runs the model
system(command_line)



##all of the code above can be copied directly into the R session in the container to run the model

```


## Run off of spinup

```{r}

# input_rhessys = IOin_rhessys_input(
#   version          = "../../rhessys7.5",
#   tec_file         = "tecfiles/tec50yr.test",
#   world_file       = "worldfiles/hpb.world.Y1999M12D31H24.state.Y1999M12D31H24.state.Y1999M12D31H24.state.Y1999M12D31H24.state.Y2019M12D31H24.state",
#   world_hdr_prefix = "hpb",
#   flowtable        = "flowtables/hpb.flow",
#   start            = "1950 01 1 1",
#   end              = "2020 01 1 1",
#   output_folder    = "out/",
#   output_prefix    = "HPB_spinup50yr_r6",
#   # commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9"
#   # swap to below for neutral spinup params:
#   commandline_options = "-b -g -s 1 10 -gw 0 0"
# )
# 
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files = "worldfiles/hpb.hdr",
#   # tec_data  = NULL,   def_pars = NULL,  output_filter = NULL,
#   return_cmd    = F   # flip to FALSE to run; T prints the command line structure
# )

```



## try a 400 year spinup with climate repeat
turn on climrepeat and off vegmort
```{r}

# input_rhessys = IOin_rhessys_input(
#   version          = "../../rhessys7.5",
#   tec_file         = "tecfiles/tec400yr.test",
#   world_file       = "worldfiles/hpb.world.Y2349M12D31H24.state",
#   world_hdr_prefix = "hpb",
#   flowtable        = "flowtables/hpb.flow",
#   start            = "1950 01 1 1",
#   end              = "2150 01 1 1",
#   output_folder    = "out/",
#   output_prefix    = "HPB_spinup400yr_plus200",
#   # commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9"
#   # swap to below for neutral spinup params:
#   commandline_options = "-b -g -s 1 10 -gw 0 0 -climrepeat"  #can add vmort_off here (check thats the right abrev)
# )
# 
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files = "worldfiles/hpb.hdr",
#   # tec_data  = NULL,   def_pars = NULL,  output_filter = NULL,
#   return_cmd    = F   # flip to FALSE to run; T prints the command line structure
# )

```



##Previous way to look at outputs

```{r}

# hpb_test1 = readin_rhessys_output("out/HPB_spinup350yr_multi_run1")
# 
# output1 <- left_join(hpb_test1$bd, hpb_test1$bdg, 
#                     by = c("day", "month", "year", "basinID", "wy", "yd", "date")) 
# 
# 
# output1 |>
# #filter(date > ymd("1980-01-01")) |>
#  # filter(date > ymd("2340-01-01")) |>
#   select(date, year, yd, 
#          lai.x, plantc.x, soilc, plantn, 
#          streamflow_DON, streamflow_DOC, streamflow_NO3,
#          streamflow, unsat_stor
#          ) |>
#   pivot_longer(-c(1:3)) |>
#   ggplot(aes(x = date, y = value))+
#   geom_point()+ labs(title = "run1")+
#   facet_wrap(~name, scales = "free_y", ncol = 3)



# ## Biogeochem Plots
# output |>
# #filter(date > ymd("1980-01-01")) |>
#  # filter(date > ymd("2020-01-01")) |>
#   select(date, year, yd, lai.x, #surfaceN,
#          litrc.x, plantc.x, soilc, plantn, soiln,
#          streamflow_DON, streamflow_DOC, streamflow_NO3, streamflow_NH4
#          ) |>
#   pivot_longer(-c(1:3)) |>   ggplot(aes(x = date, y = value))+
#   geom_point()+   facet_wrap(~name, scales = "free_y", ncol = 2)
 
# ## Hydro plots
# output |> 
#    #filter(date > ymd("1980-01-01")) |> 
#    # filter(date > ymd("2020-01-01")) |>
#   # select(date, year, yd, streamflow, baseflow, precip, recharge, unsat_stor, 
#   #        rz_storage, sat_def, 
#   #        litter_store, litter_capacity, 
#   #        evap, evap_can, evap_lit, evap_soil,
#   #        gw.Qout, gw.storage
#   #        ) |> 
#   select(date, year, yd, 
#          streamflow,  precip, recharge, unsat_stor, evap, gw.storage) |> 
#   pivot_longer(-c(1:3)) |>    ggplot(aes(x = date, y = value))+
#   geom_point()+  facet_wrap(~name, scales = "free_y", ncol = 2)

```




