---
title: "HPB_spinup_1patch"
output: html_document
date: "2026-03-15"
---

##Info on command line -s and -gw scalars
```{r}
##info on command line params: https://github.com/RHESSys/RHESSys/wiki/RHESSys-command-line-options
### -s must be followed by two parameter values for m and K
# the first value this is a multiplier on the value set for the m parameter in the soil definition files
# the second required value designates K, hydraulic conductivity at the surface - and is a multiplier on the value set for Ksat_0 in the soil definition files;

### -gw must be followed by two values for gw1 and gw2, respectively. 
#The first value is a multiplier on the sat_to_gw_coeff parameter set in the soil definition file (representing the amount of water moving from the saturated store to the groundwater store). 
#The second value is a multiplier on the gw_loss_coeff parameter in the hillslope default file (representing the amount of water moving from the groundwater store to the stream).

#defaults are: m = 0.12; K_sat0 = 0.622; sat_to_gw_coeff = 1; gw_loss_coeff = 1

```

docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop


## make worlds with one patch
This can be done in your own R outside of the docker

```{r}


# 1patch_LHS_spinup.R
# Extract one representative patch from hpb.world and run LHS parameter
# sampling on it — much faster than full basin runs for parameter screening

library(RHESSysIOinR)
library(RHESSysPreprocessing)
library(rhutils)
library(data.table)
library(lhs)
library(tidyverse)

setwd("repo/HPB_files_NEW")
options(scipen = 999)
getwd()
# -------------------- Setup output dirs --------------------
dest = "1patch_spin"
if (!dir.exists(file.path("worldfiles", dest))) dir.create(file.path("worldfiles", dest))
if (!dir.exists(file.path("flowtables", dest))) dir.create(file.path("flowtables", dest))
if (!dir.exists(file.path("out", dest))) dir.create(file.path("out", dest))

# -------------------- Read and annotate worldfile --------------------
world_path = "worldfiles/hpb.world"

world = as.data.table(read_world(world_path, hill_col = TRUE, 
                                              zone_col = TRUE, 
                                              patch_col = TRUE))


### Find veg IDs 
# Check what the actual column names are after reading
message("Columns in world: ", paste(names(world), collapse = ", "))

# Look at what's around veg_parm_ID in the raw worldfile
world[vars == "veg_parm_ID"]

# The veg parm ID value lives in the 'values' column where vars == "veg_parm_ID"
veg_ids = unique(world$values[world$vars == "veg_parm_ID"])
veg_ids = veg_ids[!is.na(veg_ids)]
message("Veg parm IDs: ", paste(veg_ids, collapse = ", "))


# Find the row index of the first veg_parm_ID entry
# unique_ID at that row gives us the patch to extract
veg_id_use = veg_ids[1]


#Get the stratum-level unique_ID for the first patch with this veg type
stratum_uid = world[level == "canopy_strata" & 
                    shift(values, 1L) == veg_id_use, unique_ID][1]

message("Stratum unique_ID: ", stratum_uid)

# Or more reliably — find the patch_ID that contains this veg type
# then get any unique_ID within that patch
target_patch_ID = world[vars == "veg_parm_ID" & values == veg_id_use, patch_ID][1]
message("Target patch_ID: ", target_patch_ID)

# Get unique_ID of any row in that patch at stratum level
extract_uid = world[patch_ID == target_patch_ID & level == "canopy_strata", unique_ID][1]
message("Extract unique_ID: ", extract_uid)

patch_world = extract_world(world, extract_uid)


# Write single patch worldfile and flowtable
patch_world_path = file.path("worldfiles", dest, 
                              paste0("hpb_1patch_vegID", veg_id_use, ".world"))
patch_flow_path  = file.path("flowtables", dest, 
                              paste0("hpb_1patch_vegID", veg_id_use, ".flow"))

write_world(patch_world, patch_world_path)

# make_1pflow generates a minimal self-draining flowtable for a single patch
patch_flow = make_1pflow(patch_world)
writeLines(patch_flow, patch_flow_path)

# Verify files were written
message("Patch worldfile: ", patch_world_path)
message("Patch flowtable: ", patch_flow_path)
file.exists(c(patch_world_path, patch_flow_path))



```


#### Run LHS sampling across 1 patch world


```{r}

library(lhs)          # install.packages("lhs")
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")


# -------------------- LHS parameter sampling --------------------
# param_ranges = list(
#   epc.root_distrib_parm = c(1,     20),
#   epc.alloc_stemc_leafc = c(1.0,   4.0),
#   epc.livewood_turnover = c(0.1,   1.0),
#   epc.froot_turnover    = c(0.1,   1.0),
#   epc.branch_turnover   = c(0.03,  0.15),
#   epc.leaf_turnover     = c(0.1,   1.0),
#   DOM_decay_rate        = c(0.001, 0.1)
# )

param_ranges = list(
  # P0 from will
  epc.root_distrib_parm = c(6,    20),   # 8 is veg decid default; 19 is github
  epc.min_percent_leafg = c(0,  2),          # def = 1; git says 0.01 to 0.001
  epc.branch_turnover   = c(0.03, 0.04),   # 0 to 1
  epc.froot_turnover    = c(0.6,  0.9),    # def = 1; 0 to 1
  epc.leaf_turnover     = c(0.1,   1.0),  # def = 1; 0 to 1
  epc.livewood_turnover = c(0.15, 0.35),  #def = 0.7; 0 to 1
  epc.gl_c              = c(0.00004, 0.00008),      #def is 0.00006; no range given.         leafscale cuticular conductance
  epc.gl_smax          = c(0.004, 0.009), # def is 0.006;  no range given.         stomatal conductance → more CO2 uptake
  epc.proj_sla         = c(1,    60),    # def is 32 in decid; 9 is git default specific leaf area → more leaf area per C
  #DWH added
  epc.alloc_stemc_leafc = c(1, 5),  #def = 2.2; git = 1.62
  DOM_decay_rate        = c(0.03, 0.06),
  epc.flnr             = c(0.05,  0.3),   # def and git say just 0.1 Rubisco efficiency → more C fixation
  epc.alloc_frootc_leafc = c(0.8, 1.6) # 1.2 and 1.325 by def and git less to fine roots = more structural C
)

n_params = length(param_ranges)
n_runs   = 96   # can push to 100+ since 1patch runs are very fast; have 14 cores on my computer

set.seed(42)
lhs_raw    = randomLHS(n_runs, n_params)
lhs_scaled = as.data.frame(lhs_raw)
colnames(lhs_scaled) = names(param_ranges)

for (p in names(param_ranges)) {
  lo = param_ranges[[p]][1]
  hi = param_ranges[[p]][2]
  lhs_scaled[[p]] = lo + lhs_scaled[[p]] * (hi - lo)
}

message("LHS design: ", n_runs, " runs x ", n_params, " parameters")
print(summary(lhs_scaled))

# -------------------- Build inputs --------------------
# Single patch runs are very fast so can run long — 250 years with climrepeat
dates = c("1950 1 1 1", "2200 1 2 24")
name  = "hpb_1patch_lhs"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = paste0("tecfiles/", name, ".tec"),
  world_file       = "worldfiles/1patch_spin_KEEP/hpb_1patch_vegID3.world",
  world_hdr_prefix = name,
  flowtable        = "flowtables/1patch_spin/hpb_1patch_vegID3.flow",
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/1patch_spin/LHS_trial3",
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat"  # -vmort_off    -climrepeat
)

input_hdr = IOin_hdr(
  basin        = "defs/basin.def",
  hillslope    = "defs/hill.def",
  zone         = "defs/zone.def",
  soil         = "defs/soil_siltyloam.def",
  landuse      = "defs/lu_undev.def",
  stratum      = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], 
                               output_state = FALSE)

# -------------------- Pass full LHS vectors to run_rhessys_multi --------------------
pars_list = list(
  ## set hydro parameters
  list(input_hdr$soil_def[1], "m", 0.108),
  list(input_hdr$soil_def[1], "Ksat_0",  404.3),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  0.1),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff", 0.9),

  list(input_hdr$stratum_def[1], "epc.root_distrib_parm", lhs_scaled$epc.root_distrib_parm),
  list(input_hdr$stratum_def[1], "epc.min_percent_leafg", lhs_scaled$epc.min_percent_leafg),
  list(input_hdr$stratum_def[1], "epc.branch_turnover",    lhs_scaled$epc.branch_turnover),
  list(input_hdr$stratum_def[1], "epc.froot_turnover",     lhs_scaled$epc.froot_turnover),
  list(input_hdr$stratum_def[1], "epc.leaf_turnover",      lhs_scaled$epc.leaf_turnover),
  list(input_hdr$stratum_def[1], "epc.livewood_turnover",  lhs_scaled$epc.livewood_turnover),
  list(input_hdr$stratum_def[1],  "epc.gl_c",                lhs_scaled$epc.gl_c),
  list(input_hdr$stratum_def[1],    "epc.gl_smax",          lhs_scaled$epc.gl_smax),
  list(input_hdr$stratum_def[1],    "epc.proj_sla",          lhs_scaled$epc.proj_sla),
  
  list(input_hdr$stratum_def[1], "epc.alloc_stemc_leafc",   lhs_scaled$epc.alloc_stemc_leafc),
  list(input_hdr$soil_def[1],    "DOM_decay_rate",          lhs_scaled$DOM_decay_rate),
  list(input_hdr$stratum_def[1],  "epc.flnr",              lhs_scaled$epc.flnr),
  list(input_hdr$stratum_def[1], "epc.alloc_frootc_leafc",  lhs_scaled$epc.alloc_frootc_leafc)
  )


input_def_pars = IOin_def_pars_simple(pars_list)

# -------------------- Run --------------------
run_rhessys_multi(
  input_rhessys    = input_rhessys,
  hdr_files        = input_hdr,
  tec_data         = input_tec_data,
  def_pars         = input_def_pars,
  output_filter    = NULL,
  all_combinations = FALSE,
  return_cmd       = F   # flip to FALSE to run
)

# Save parameter table
lhs_scaled |>
  mutate(run = paste0(name, "_run", seq_len(n_runs))) |>
  write_csv("out/1patch_spin/LHS_trial3/lhs_1patch_parameter_table.csv")


```






evaluate

```{r}

# -------------------- Evaluate results --------------------
# Reuse the same evaluation code as before, just point at the 1patch output folder

target_plantc = 15
target_soilc  = 15

out_files = list.files("out/1patch_spin/LHS_trial3", pattern = "hpb_1patch_lhs_run.*basin\\.daily$", 
                        full.names = TRUE)
out_files = out_files[!grepl("grow", out_files)]  # exclude the grow files

all_ts = map_dfr(out_files, function(f) {
  run_prefix = gsub("_basin\\.daily$", "", f)
  run        = gsub(".*/", "", run_prefix)
  message("Reading: ", run)
  
  dat = tryCatch(
    readin_rhessys_output(run_prefix),
    error = function(e) { message("Failed: ", e$message); return(NULL) }
  )
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  out |>
    select(date, plantc = plantc.x, soilc, lai = lai.x, streamflow) |>
    mutate(run = run)
})


## Read in param table
param_table = read_csv("out/1patch_spin/LHS_trial3/lhs_1patch_parameter_table.csv")

### Results sumary
results = all_ts |>
  group_by(run) |>
  slice_tail(n = 365) |>
  summarise(
    mean_plantc     = max(plantc,     na.rm = TRUE),
    mean_soilc      = max(soilc,      na.rm = TRUE),
    mean_lai        = max(lai,         na.rm = TRUE),
    mean_streamflow = max(streamflow,  na.rm = TRUE)
  ) |>
  left_join(param_table, by = "run") |>
  # mutate(score = abs(mean_plantc - target_plantc) + 
  #                abs(mean_soilc  - target_soilc)) |>
    mutate(score = abs(mean_plantc - mean_soilc)) |>
  arrange(score)

print(head(results, 10))

# -------------------- Time series plot — all runs overlaid --------------------
# Color by score so best runs stand out
all_ts |>
  #filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
  mutate(Cstock_dif = plantc - soilc) |> 
  pivot_longer(cols = c(plantc, soilc, lai, streamflow, Cstock_dif),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")



# Top 5 trajectories
top_runs = results$run[1:5]

all_ts |>
  filter(run %in% top_runs) |>
  left_join(results |> select(run, score), by = "run") |>
 mutate(Cstock_dif = plantc - soilc) |> 
  pivot_longer(cols = c(plantc, soilc, lai, streamflow, Cstock_dif),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, color = run)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "Top 5 single-patch LHS runs",
       x = "Date", y = NULL) +
  theme_bw()



```







#### TRYING WORKFLOW ABOVE BUT RUN BY parameter 
NOTE READ!!!!!! This currently runs 36 iterations across 13 parameters; this makes a lot of files and filled up my one drive, also super slow read in data..
Consider doing less variables and iterations next time..

```{r}
library(lhs)
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")

# -------------------- Fixed hydro parameters (same for all runs) --------------------
fixed_m               = 0.108
fixed_Ksat_0          = 404.3
fixed_sat_to_gw_coeff = 0.1
fixed_gw_loss_coeff   = 0.9

# -------------------- Default/fixed veg parameter values --------------------
# These are held constant while each other parameter varies in turn
fixed_veg_pars = list(
  epc.root_distrib_parm  = 8,
  epc.min_percent_leafg  = 1,
  epc.branch_turnover    = 0.035, #0
  epc.froot_turnover     = 0.75, #1
  epc.leaf_turnover      = 0.5, #1
  epc.livewood_turnover  = 0.25, #0.7
  epc.gl_c               = 0.00006,
  epc.gl_smax            = 0.006,
  epc.proj_sla           = 32,
  epc.alloc_stemc_leafc  = 2.2,
  DOM_decay_rate         = 0.045, #0.05
  epc.flnr               = 0.1,
  epc.alloc_frootc_leafc = 1.2
)

# -------------------- Parameter ranges to sweep one at a time --------------------
#could also prob set this as c(super low, low, mid, high, super high) then just have to do 5 per. similar to rheyssy multi in first script 
param_ranges = list(
  epc.root_distrib_parm  = c(6,     20),
  epc.min_percent_leafg  = c(0,     2),
  epc.branch_turnover    = c(0.03,  0.04),
  epc.froot_turnover     = c(0.6,   0.9),
  epc.leaf_turnover      = c(0.1,   1.0),
  epc.livewood_turnover  = c(0.15,  0.35),
  epc.gl_c               = c(0.00004, 0.00008),
  epc.gl_smax            = c(0.004,  0.009),
  epc.proj_sla           = c(1,     60),
  epc.alloc_stemc_leafc  = c(1,     5),
  DOM_decay_rate         = c(0.03,  0.06),
  epc.flnr               = c(0.05,  0.3),
  epc.alloc_frootc_leafc = c(0.8,   1.6)
)

n_runs   = 36
dates    = c("1950 1 1 1", "2200 1 2 24")

# -------------------- Loop over each parameter --------------------
for (focal_param in names(param_ranges)) {
  
  message("\n========================================")
  message("Running parameter: ", focal_param)
  message("========================================")
  
  # Create output folder for this parameter
  out_folder = file.path("out/1patch_spin/LHS_byparam", focal_param)
  if (!dir.exists(out_folder)) dir.create(out_folder, recursive = TRUE)
  
  run_name = paste0("hpb_1p_", focal_param)
  
  # -------------------- Generate LHS for focal parameter only --------------------
  set.seed(42)
  lo = param_ranges[[focal_param]][1]
  hi = param_ranges[[focal_param]][2]
  
  # LHS over single dimension — evenly spaced samples across the range
  lhs_vals = lo + randomLHS(n_runs, 1) * (hi - lo)
  lhs_vals = as.vector(lhs_vals)
  
  # Save parameter table for this focal param
  param_table = tibble(
    run           = paste0(run_name, "_run", seq_len(n_runs)),
    focal_param   = focal_param,
    focal_value   = lhs_vals
  )
  # Also record all fixed values for full reproducibility
  for (fp in names(fixed_veg_pars)) {
    param_table[[fp]] = ifelse(fp == focal_param, NA, fixed_veg_pars[[fp]])
  }
  param_table[[focal_param]] = lhs_vals
  
  write_csv(param_table, file.path(out_folder, "parameter_table.csv"))
  
  # -------------------- Build inputs --------------------
  input_rhessys = IOin_rhessys_input(
    version          = "../../rhessys7.5",
    tec_file         = paste0("tecfiles/", run_name, ".tec"),
    world_file       = "worldfiles/1patch_spin_KEEP/hpb_1patch_vegID3.world",
    world_hdr_prefix = run_name,
    flowtable        = "flowtables/1patch_spin/hpb_1patch_vegID3.flow",
    start            = dates[1],
    end              = dates[2],
    output_folder    = paste0(out_folder, "/"),
    output_prefix    = run_name,
    commandline_options = "-b -g -climrepeat"
  )
  
  input_hdr = IOin_hdr(
    basin        = "defs/basin.def",
    hillslope    = "defs/hill.def",
    zone         = "defs/zone.def",
    soil         = "defs/soil_siltyloam.def",
    landuse      = "defs/lu_undev.def",
    stratum      = "defs/veg_deciduous.def",
    basestations = "clim/ccr_base"
  )
  
  input_tec_data = IOin_tec_std(start = dates[1], end = dates[2],
                                 output_state = FALSE)
  
  # -------------------- Build pars_list --------------------
  # Fixed hydro params first
  pars_list = list(
    list(input_hdr$soil_def[1],      "m",                fixed_m),
    list(input_hdr$soil_def[1],      "Ksat_0",           fixed_Ksat_0),
    list(input_hdr$soil_def[1],      "sat_to_gw_coeff",  fixed_sat_to_gw_coeff),
    list(input_hdr$hillslope_def[1], "gw_loss_coeff",    fixed_gw_loss_coeff)
  )
  
  # All veg/soil params — focal one gets LHS vector, others get fixed scalar
  all_veg_params = names(fixed_veg_pars)
  
  for (p in all_veg_params) {
    
    # Determine which def file this parameter belongs to
    def_file = if (p == "DOM_decay_rate") {
      input_hdr$soil_def[1]
    } else {
      input_hdr$stratum_def[1]
    }
    
    # Focal param gets the LHS vector, all others get their fixed scalar
    val = if (p == focal_param) lhs_vals else fixed_veg_pars[[p]]
    
    pars_list = c(pars_list, list(list(def_file, p, val)))
  }
  
  input_def_pars = IOin_def_pars_simple(pars_list)
  
  # -------------------- Run --------------------
  run_rhessys_multi(
    input_rhessys    = input_rhessys,
    hdr_files        = input_hdr,
    tec_data         = input_tec_data,
    def_pars         = input_def_pars,
    output_filter    = NULL,
    all_combinations = FALSE,
    return_cmd       = FALSE
  )
  
  message("Completed: ", focal_param)
}

message("\nAll parameter sweeps complete!")
message("Results in: out/1patch_spin/LHS_byparam/")

```

The folder structure this produces:
```
out/1patch_spin/LHS_byparam/
  epc.root_distrib_parm/
    parameter_table.csv
    hpb_1p_epc.root_distrib_parm_run1_basin.daily
    hpb_1p_epc.root_distrib_parm_run2_basin.daily
    ...
  epc.min_percent_leafg/
    parameter_table.csv
    ...
  epc.froot_turnover/
    ...
```


EVALUATE

```{r}
# -------------------- Read results by parameter --------------------

# Store all dataframes in a named list — one per parameter
all_ts_by_param = list()

for (focal_param in names(param_ranges)) {
  
  message("Reading: ", focal_param)
  
  out_folder  = file.path("out/1patch_spin/LHS_byparam", focal_param)
  param_table = read_csv(file.path(out_folder, "parameter_table.csv"),
                          show_col_types = FALSE)
  
  out_files = list.files(out_folder, pattern = "basin\\.daily$", full.names = TRUE)
  out_files = out_files[!grepl("grow", out_files)]
  
  ts = map_dfr(out_files, function(f) {
    run_prefix = gsub("_basin\\.daily$", "", f)
    run        = gsub(".*/", "", run_prefix)
    
    dat = tryCatch(
      readin_rhessys_output(run_prefix),
      error = function(e) { message("  Failed: ", e$message); return(NULL) }
    )
    if (is.null(dat)) return(NULL)
    
    out = left_join(dat$bd, dat$bdg,
                    by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
    
    out |>
      select(date, year, month, day,
             plantc = plantc.x, soilc, lai = lai.x, streamflow) |>
      mutate(run = run)
  })
  
  # Join with parameter values so each row knows its focal param value
  ts = ts |>
    left_join(param_table |> select(run, focal_value), by = "run")
  
  all_ts_by_param[[focal_param]] = ts
  
  message("  Done — ", nrow(ts), " rows, ", length(out_files), " runs")
}

# -------------------- Access individual dataframes --------------------
# # Each parameter's data is now a named element of the list, e.g.:
# all_ts_by_param[["epc.root_distrib_parm"]]
# all_ts_by_param[["epc.froot_turnover"]]
# 
# # Or assign to individual named dataframes if you prefer working with them directly
# list2env(all_ts_by_param, envir = .GlobalEnv)
# This creates objects named exactly after each parameter in your environment:
# epc.root_distrib_parm, epc.froot_turnover, etc.
```



```{r}
# -------------------- Plot one parameter --------------------
plot_param_ts = function(param_name) {
  all_ts_by_param[[param_name]] |>
    filter(date > ymd("2050-01-01")) |>
    pivot_longer(cols = c(plantc, soilc, lai, streamflow),
                 names_to = "variable", values_to = "value") |>
    ggplot(aes(x = as.Date(date), y = value, 
               group = run, color = focal_value)) +
    geom_line(alpha = 0.6, linewidth = 0.3) +
    scale_color_viridis_c(name = param_name) +
    facet_wrap(~variable, scales = "free_y", ncol = 2) +
    labs(title = paste("Sensitivity:", param_name),
         x = "Date", y = NULL) +
    theme_bw()
}

# Single parameter
plot_param_ts("epc.root_distrib_parm")
plot_param_ts("epc.min_percent_leafg")
plot_param_ts("epc.branch_turnover")
plot_param_ts("epc.froot_turnover")
plot_param_ts("epc.leaf_turnover")
plot_param_ts("epc.livewood_turnover")
plot_param_ts("epc.gl_c")
plot_param_ts("epc.gl_smax")
plot_param_ts("epc.proj_sla")
plot_param_ts("epc.alloc_stemc_leafc")
plot_param_ts("DOM_decay_rate")
plot_param_ts("epc.flnr")
plot_param_ts("epc.alloc_frootc_leafc")



# -------------------- Loop to plot all and save to PDF --------------------
# pdf("out/1patch_spin/LHS_byparam/sensitivity_plots.pdf", width = 10, height = 8)
# for (p in names(param_ranges)) {
#   plot_param_ts(p)
# }
# dev.off()

```









