---
title: "HPB_spinup"
output: html_document
date: "2026-03-12"
---

##Info on command line -s and -gw scalars
```{r}
##info on command line params: https://github.com/RHESSys/RHESSys/wiki/RHESSys-command-line-options
### -s must be followed by two parameter values for m and K
# the first value this is a multiplier on the value set for the m parameter in the soil definition files
# the second required value designates K, hydraulic conductivity at the surface - and is a multiplier on the value set for Ksat_0 in the soil definition files;

### -gw must be followed by two values for gw1 and gw2, respectively. 
#The first value is a multiplier on the sat_to_gw_coeff parameter set in the soil definition file (representing the amount of water moving from the saturated store to the groundwater store). 
#The second value is a multiplier on the gw_loss_coeff parameter in the hillslope default file (representing the amount of water moving from the groundwater store to the stream).

#defaults are: m = 0.12; K_sat0 = 0.622; sat_to_gw_coeff = 1; gw_loss_coeff = 1

```

docker run -v C:/Users/dwh18/OneDrive/Desktop/R_projects/RHESSys_Tutorial:/RHESSys/repo -it rhessys/rhessys:develop


Guidance from spinup on what variables to evaluate
To determine if the model has stabilized, you are interested in looking at each of the following output variables from the grow_basin.daily files: plantc, plantn, soilc, soiln. First look at output after 75 years of processing, then after 225 years of processing in order to see how these variables have changed and moved toward stabilization.
You may also want to look at streamflow, LAI, and soil moisture (computed by subtracting the variable unsat_stor from sat_def)


## run using run_rhessys command 
this is for HPB_files_NEW folder that uses the kmeans500 patches 

```{r}
#Can run all of this within the Docker container after opening R

setwd("repo/HPB_files_NEW")
library(RHESSysIOinR)

dates = c("1950 1 1 1", "2100 1 2 24") #has to end with a 24 hour so that the tec output will work

name = "hpb_spinup_Hydro"

## possible way to pull best world file
# Use your best current .state file as starting point
# worldfile_start = tail(list.files("worldfiles", pattern = "\\.state", full.names = TRUE), 1)
# message("Starting from: ", worldfile_start) ## would then set world)file in input rhessys to worldfile_start

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         =  paste0("tecfiles/",name,".tec"), #for using tecfile built below
  # tec_file         = "tecfiles/tec50yr.test", #call local tecfile
  world_file       = "worldfiles/hpb.world", #what was made in pre-processing
  world_hdr_prefix = "hpb_spintest",
  flowtable        = "flowtables/hpb.flow", #what was made in pre-processing
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/HydroTests/",
  output_prefix    = "HPB_spinup150yr_HydroX",
  # commandline_options = "-b -g -s 0.9 650 -gw 0.1 0.9 -climrepeat"
  #commandline_options = "-b -g -s 1 100 -gw 1 1"
  # swap to below for neutral spinup params: 
  # commandline_options = "-b -g -s 1 10 -gw 0 0 -climrepeat"
  commandline_options = "-b -g -climrepeat"  #having these set as nothing seems to not let anything grow
)

#header directly copied from hpb.hdr
input_hdr = IOin_hdr(
  basin      = "defs/basin.def",
  hillslope  = "defs/hill.def",
  zone       = "defs/zone.def",
  soil       = "defs/soil_siltyloam.def",
  landuse    = "defs/lu_undev.def",
  stratum    = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)


#where you can manually tweak params
# pars_list = list(
#   # list(def_file_path,       "parameter_name",    value) 
#   list(input_hdr$stratum_def[1], "epc.root_distrib_parm", 18), #default is 8 in deciduous
#   list(input_hdr$stratum_def[1], "epc.leaf_turnover", 1), #default is 1 in deciduous; range is 0-1: %  leaves that die each year
#   list(input_hdr$stratum_def[1], "epc.livewood_turnover", 0.7) #default is 0.7 in deciduous
#   # add as many as you need in the same format
# )

pars_list = list(
  list(input_hdr$soil_def[1],            "m",    c(0.12,   0.108,    1    )),      #0.12
  list(input_hdr$soil_def[1],          "Ksat_0", c(0.622,  404.3,    10   )),     #0.622
  list(input_hdr$soil_def[1], "sat_to_gw_coeff", c(1,       0.1,      0   )),   #1
  list(input_hdr$hillslope[1], "gw_loss_coeff",  c(1,       0.9,      0   ))     #1
  )



# #set up manually tweak params for rhessys format; can also put multi here  #when you run this it will generate new folders within def called zone, soil, etc, that have updated def file values, but doesn't change original files
input_def_pars = IOin_def_pars_simple(pars_list)



#make tec file locally
input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], output_state = T)



#run model with inputs defined above
# run_rhessys_single(
#   input_rhessys = input_rhessys,
#   hdr_files     = input_hdr,
#   tec_data      = input_tec_data, #set in input rhessys; can be NULL if calling a local file and not using IOin_tec_std
#   # def_pars      = input_def_pars,
#   output_filter = NULL, #cant get any version working currently so using legacy output
#   return_cmd    = F   # flip to FALSE to run
# )


#run for multi param
run_rhessys_multi(
  input_rhessys   = input_rhessys,
  hdr_files       = input_hdr,
  tec_data        = input_tec_data,
  def_pars        = input_def_pars,
  output_filter   = NULL,
  all_combinations = FALSE,   # paired not grid; grid would run all combos of inputs list
  return_cmd      = F
)

##seems like it runs by pair for each year: ie 57,57,57, 58,58,58 . at some point it seems to get off tho.. ie 2272, 2259, 2237
## started ~632 - 657




```



## look at output


```{r}
##functin to read in and plot quickly

library(RHESSysIOinR)
library(tidyverse)


read_plot_rhessys = function(outname, date_filter = NULL, ncol = 3,
                              vars = c("lai.x", "plantc.x", "soilc", "plantn", "soiln",
                                       "streamflow", "unsat_stor", "sat_def",
                                       "streamflow_DON", "streamflow_DOC", "streamflow_NO3")) {
  
  dat = readin_rhessys_output(outname)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  if (!is.null(date_filter)) out = out |> filter(date > date_filter)
  
  out |>
    select(date, year, yd, any_of(vars)) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    pivot_longer(-c(date, year, yd)) |>
    ggplot(aes(x = date, y = value)) +
    geom_point() +
    facet_wrap(~name, scales = "free_y", ncol = ncol) +
    labs(title = outname, x = "Date", y = NULL) +
    theme_bw()
}

# read_plot_rhessys(outname = "out/Trial1/HPB_spinup350yr_multi_run1", date_filter = NULL)


```




```{r}

read_plot_rhessys(outname = "out/HydroTests/Hydro1/HPB_spinup150yr_Hydro1", date_filter = ymd("1970-01-01") ) # NULL is default ymd("2000-01-01")

read_plot_rhessys(outname = "out/HydroTests/Hydro2/HPB_spinup150yr_Hydro2", date_filter = ymd("1970-01-01") )

read_plot_rhessys(outname = "out/HydroTests/Hydro3_5/HPB_spinup150yr_HydroX_run1", date_filter = ymd("1970-01-01") )

read_plot_rhessys(outname = "out/HydroTests/Hydro3_5/HPB_spinup150yr_HydroX_run2", date_filter = ymd("1970-01-01") )

read_plot_rhessys(outname = "out/HydroTests/Hydro3_5/HPB_spinup150yr_HydroX_run3", date_filter = ymd("1970-01-01") )

```



## LHS sampling across the hydro params 


```{r}
library(lhs)          # install.packages("lhs")
library(RHESSysIOinR)
library(tidyverse)
library(data.table)

setwd("repo/HPB_files_NEW")


# -------------------- Define parameter ranges --------------------
##values from testing above
# pars_list = list(
#   list(input_hdr$soil_def[1],            "m",    c(0.12,   0.108,    1    )),      #0.12
#   list(input_hdr$soil_def[1],          "Ksat_0", c(0.622,  404.3,    10   )),     #0.622
#   list(input_hdr$soil_def[1], "sat_to_gw_coeff", c(1,       0.1,      0   )),   #1
#   list(input_hdr$hillslope[1], "gw_loss_coeff",  c(1,       0.9,      0   ))     #1
#   )

#ranges from spinup instructtions
param_ranges = list(
  m                  = c(0.05,     2),  #0.12
  Ksat_0            = c(1,   600),  # 0.622
  sat_to_gw_coeff   = c(0.001,   1), # 1
  gw_loss_coeff     = c(0.01,   1)  # 1
)

n_params = length(param_ranges)
n_runs   = 12

# -------------------- Generate LHS design --------------------
set.seed(42)
lhs_raw    = randomLHS(n_runs, n_params)
lhs_scaled = as.data.frame(lhs_raw)
colnames(lhs_scaled) = names(param_ranges)

for (p in names(param_ranges)) {
  lo = param_ranges[[p]][1]
  hi = param_ranges[[p]][2]
  lhs_scaled[[p]] = lo + lhs_scaled[[p]] * (hi - lo)
}

print(head(lhs_scaled))
print(summary(lhs_scaled))

# -------------------- Setup --------------------
dates          = c("1950 1 1 1", "2050 1 2 24")
name           = "hpb_HydroLHS2"
worldfile_start = "worldfiles/hpb.world"

input_rhessys = IOin_rhessys_input(
  version          = "../../rhessys7.5",
  tec_file         = paste0("tecfiles/", name, ".tec"),
  world_file       = worldfile_start,
  world_hdr_prefix = name,
  flowtable        = "flowtables/hpb.flow",
  start            = dates[1],
  end              = dates[2],
  output_folder    = "out/HydroTests/HydroLHS2/",
  output_prefix    = name,
  commandline_options = "-b -g -climrepeat"
)

input_hdr = IOin_hdr(
  basin        = "defs/basin.def",
  hillslope    = "defs/hill.def",
  zone         = "defs/zone.def",
  soil         = "defs/soil_siltyloam.def",
  landuse      = "defs/lu_undev.def",
  stratum      = "defs/veg_deciduous.def",
  basestations = "clim/ccr_base"
)

input_tec_data = IOin_tec_std(start = dates[1], end = dates[2], 
                               output_state = FALSE)

# -------------------- Pass full vectors to run_rhessys_multi --------------------
# No [i] indexing — pass the entire column as a vector
pars_list = list(
  list(input_hdr$soil_def[1], "m", lhs_scaled$m),
  list(input_hdr$soil_def[1], "Ksat_0",  lhs_scaled$Ksat_0),
  list(input_hdr$soil_def[1], "sat_to_gw_coeff",  lhs_scaled$sat_to_gw_coeff),
  list(input_hdr$hillslope_def[1], "gw_loss_coeff",     lhs_scaled$gw_loss_coeff)
)

input_def_pars = IOin_def_pars_simple(pars_list)

# -------------------- Run --------------------
run_rhessys_multi(
  input_rhessys    = input_rhessys,
  hdr_files        = input_hdr,
  tec_data         = input_tec_data,
  def_pars         = input_def_pars,
  output_filter    = NULL,
  all_combinations = FALSE,
  return_cmd       = F   # flip to FALSE to run
)

# -------------------- Save param table --------------------
lhs_scaled |>
  mutate(run = paste0(name, "_run", seq_len(n_runs))) |>
  write_csv("out/HydroTests/HydroLHS2/lhs_parameter_table.csv")




```









evaluate LHS

```{r}
library(tidyverse)
library(data.table)
library(RHESSysIOinR)


#### UPDATE FILE PATH
out_files = list.files("out/HydroTests/HydroLHS2", pattern = "hpb_HydroLHS2_run.*_basin\\.daily$", 
                        full.names = TRUE)
out_files = out_files[!grepl("grow", out_files)]  # exclude the grow files

all_ts = map_dfr(out_files, function(f) {
  
  # Base prefix is everything before _basin.daily
  # e.g. out/hpb_lhs_run001_basin.daily -> out/hpb_lhs_run001
  run_prefix = gsub("_basin\\.daily$", "", f)
  run        = gsub(".*/", "", run_prefix)
  
  message("Reading: ", run)
  
  dat = tryCatch(
    readin_rhessys_output(run_prefix),
    error = function(e) { 
      message("  Failed: ", e$message)
      return(NULL) 
    }
  )
  
  if (is.null(dat)) return(NULL)
  
  out = left_join(dat$bd, dat$bdg,
                  by = c("day", "month", "year", "basinID", "wy", "yd", "date"))
  
  out |>
    select(date, year, month, day,
           plantc = plantc.x,
           soilc,
           plantn,
           soiln,
           streamflow_DOC,
           streamflow_DON,
           streamflow_NO3,
           lai    = lai.x,
           unsat_stor, 
           sat_def,
           streamflow) |>
    mutate(soil_moisture =  sat_def - unsat_stor) |> 
    mutate(run = run)
})


# -------------------- Read data --------------------
#### UPDATE FILE PATH
param_table = read_csv("out/HydroTests/HydroLHS2/lhs_parameter_table.csv")


# -------------------- Scoring table (unchanged) --------------------
results = all_ts |>
  group_by(run) |>
  slice_tail(n = 365) |>
  summarise(
    mean_plantc     = max(plantc,     na.rm = TRUE),
    mean_soilc      = max(soilc,      na.rm = TRUE),
    mean_lai        = max(lai,         na.rm = TRUE),
    mean_streamflow = max(streamflow,  na.rm = TRUE),
    mean_streamflow_DOC = max(streamflow_DOC,  na.rm = TRUE),
    mean_streamflow_NO3 = max(streamflow_NO3,  na.rm = TRUE)
  ) |>
  left_join(param_table, by = "run") |>
  mutate(score = abs(mean_plantc - 15) + 
                 abs(mean_soilc  - 15)) |> #using 15 as target kg/m2 for both pools curretnly
  arrange(score)

print(head(results, 10))

# -------------------- Time series plot — all runs overlaid --------------------
# Color by score so best runs stand out
all_ts |>
  filter(date > ymd("2000-01-01")) |> 
  left_join(results |> select(run, score), by = "run") |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow, streamflow_DOC, streamflow_NO3),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, 
             group = run, color = score)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_viridis_c(direction = -1, name = "Score\n(lower=better)") +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "LHS runs — full time series",
       subtitle = "Brighter lines = closer to target C stocks",
       x = "Date", y = NULL) +
  theme_bw() +
  theme(legend.position = "right")

# -------------------- Highlight just the top N runs --------------------
top_n = 5

top_runs = results$run[1:top_n]

all_ts |>
  filter(run %in% top_runs) |>
  left_join(results |> select(run, score), by = "run") |>
  mutate(run_label = paste0(run, "\nscore=", round(score, 2))) |>
  pivot_longer(cols = c(plantc, soilc, lai, streamflow_DOC, streamflow, soil_moisture),
               names_to = "variable", values_to = "value") |>
  ggplot(aes(x = as.Date(date), y = value, color = run_label)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title    = paste("Top", top_n, "LHS runs by score"),
       subtitle = "Are stocks leveling off by end of run?",
       x = "Date", y = NULL, color = "Run") +
  theme_bw() +
  theme(legend.position = "bottom")

# -------------------- Stabilization check --------------------
# Compare mean of first vs last year to see if stocks are still drifting
all_ts |>
  left_join(results |> select(run, score), by = "run") |>
  group_by(run, score) |>
  mutate(run_year = year(as.Date(date)),
         first_year = min(run_year),
         last_year  = max(run_year)) |>
  summarise(
    plantc_start = mean(plantc[run_year == first_year], na.rm = TRUE),
    plantc_end   = mean(plantc[run_year == last_year],  na.rm = TRUE),
    soilc_start  = mean(soilc[run_year  == first_year], na.rm = TRUE),
    soilc_end    = mean(soilc[run_year  == last_year],  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    plantc_drift = round(plantc_end - plantc_start, 3),
    soilc_drift  = round(soilc_end  - soilc_start,  3)
  ) |>
  arrange(score) |>
  select(run, score, plantc_start, plantc_end, plantc_drift, 
         soilc_start, soilc_end, soilc_drift) |>
  print(n = 20)


```






















































